
# Sample size evaluation {#samplesize}

Sample size calculations provide an insight in the relationship between sample size and the minimal detectable effect size for a given significance level $\alpha$ and power $\pi$. $\alpha$ is the probability of a type I error, i.e. the probability of concluding that there is an effect based on a sample, when in reality there is no effect in the population. $\pi$ is the probability to detect an effect based on a sample, when the effect also exists in reality.

The original sample size of the monitoring scheme is based on sample size calculations for the status of the habitat quality [@onkelinx2007, @westra2014]. We will first reproduce the sample size calculations by @onkelinx2007 Next we will also perform a sample size calculation for the trend in habitat quality.

As in @westra2014, we will choose a power $\pi$ of 0.80 and a significance level $\alpha$ of 0.05.    

## Habitat quality status

### Minimal detectable difference between the proportion of habitat in bad condition and the 25%-threshold

The quality of a habitat type is unfavourable (on a regional scale) when more than 25% of the habitat area is in a bad condition. Here we want to determine the minimal detectable difference from the 25%-threshold.

For a straightforward sample size calculation for the minimal detectable difference from a threshold value, we can apply the *proportion test* [@cohen1988] using the function `pwr.p.test` of the pwr package [@champely2020].

```{r hulpfunctie}

Es.h_inv <- function(h, p2) {
  (sin((h + 2*asin(sqrt(p2))) / 2))^2
}

p1 <- 0.3
p2 <- 0.25

check <- p1  -  Es.h_inv(ES.h(p1, p2), p2) < 0.001
```

```{r powerverschil}

sim_data <- expand.grid(
  n_sim = seq(20, 1000, 1),
  alpha = c(0.05),
  power = c(0.80),
  p2 = 0.25
)
sim_data$p1_low <- NULL
sim_data$p1_high <- NULL

for (i in 1:nrow(sim_data)) {
  
  h = (pwr.p.test(sig.level = sim_data$alpha[i],
                        power = sim_data$power[i],
                        alternative = "two.sided",
                        n = sim_data$n_sim[i]))$h
  
  sim_data$p1_low[i] <- round(Es.h_inv(-h, sim_data$p2[i]), 3)
  sim_data$p1_high[i] <- round(Es.h_inv(h,sim_data$p2[i]), 3)
}

```

Figure \@ref(fig:figpowerverschil) and Table \@ref(tab:tab-powerverschil) show the results of the sample size calculations. The results can be interpreted as follows. When in reality less than 16.3 % of the habitat is in bad condition, we will be able to conclude that the habitat proportion in bad condition is significantly lower than the 25%-threshold when we have sample size of 170 (with $\pi = 0.80$ and $\alpha = 0.05$). Hence, the minimal detectable difference is 8,7 %. Similarly, when in reality more than 34,8 % of the habitat has a bad condition, we will be able to conclude that the habitat proportion with a bad quality is significantly higher than the 25%-threshold. When the habitat proportion with a bad quality is between 16,3 % and 34,8 %, a sample size of 170 will not allow us to detect a difference from the 25%-threshold with a probability of at least 0.80 ($\pi$) and a type I error of at most 0.05 ($\alpha$).

```{r figpowerverschil, fig.cap = "Relationship between sample size and minimal detectable difference"}

ggplot() +
  geom_line(data = sim_data, aes(x = n_sim, y = p1_low)) +
  geom_line(data = sim_data, aes(x = n_sim, y = p1_high)) +
  geom_hline(yintercept = 0.25, linetype = 1, size = 0.5, alpha = 0.5) +
  labs(x = "Sample size", y = "Habitat propotion in bad condition") +
  scale_y_continuous(labels = scales::percent) +
  scale_linetype_manual(values = c(2,3))

```

```{r tab-powerverschil, eval =TRUE}
sim_data_tabel <- sim_data %>%
  filter(n_sim %in% c(40, 80, 170, 400, 1000)) %>%
  mutate(favourable = str_c("< ", p1_low * 100, " %"),
         unfavourable = str_c("> ", p1_high * 100, " %"),
         lower = str_c((p1_low - p2) * 100, "%"),
         higher = str_c("+",(p1_high - p2) * 100, " %")) %>%
  dplyr::select("sample size" = n_sim, favourable, unfavourable, lower, higher)

sim_data_tabel %>%
  kable(align = "c", 
        caption = "Minimal detectable difference (mdd) from 25\\% - threshold as a function of the sample size with $\\alpha = 0.05$ and $\\pi = 0.80$",
        booktabs = TRUE, 
        format = format_table) %>%
  kable_styling() %>%
  add_header_above(c(" " = 3, "mdd from 25%" = 2))
```

### Comparison with results of the 2019 habitat reporting

```{r}

status_bos <- read.csv2("data/statushabitat_bos_toestand.csv")
status_grasland_moeras <- read.csv2("data/statushabitat_graslandMoeras.csv")
statushabitat_6510 <- read.csv2("data/statushabitat_6510.csv")
statushabitat_heide <- read.csv2("data/statushabitat_heide.csv")
statushabitat_pink <- read.csv2("data/statushabitat_pink.csv")

status_types <- bind_rows(
  status_bos,
  status_grasland_moeras,
  statushabitat_6510,
  statushabitat_heide,
  statushabitat_pink
) %>%
  filter(TypeResultaat == "Habitattype",
         Schaal == "Vlaanderen",
         Versie == "Versie 3",
         nObs > 5) %>%
  dplyr::select(type = Habitattype, n = nObs, prop_goed = AandeelGunstig, prop_goed_lci = AandeelGunstig_LLCI, prop_goed_uci = AandeelGunstig_ULCI) %>%
  mutate(type = ifelse(type == "91", "91E0", as.character(type))) %>%
  arrange(type) %>%
  mutate(prop_slecht = (100 - prop_goed)/100,
         prop_slecht_lci = (100 - prop_goed_lci)/100,
         prop_slecht_uci = (100 - prop_goed_uci)/100,
         verschil_grenswaarde = round(prop_slecht - 0.25, 3),
     uitspraak  = ifelse(prop_goed_lci > 75, "Favourable",
                             ifelse(prop_goed_uci < 75, "Unfavourable", "Unknown")))

```

In Figure \@ref(fig:figrapportage) we show the estimated habitat proportions in bad condition for terrestrial habitats, as reported in 2019. The figure also shows the 95% - confidence interval of the estimate and the sample size on which the estimate is based (text above the confidence interval). The estimates are ordered in the x-axis from low to high sample size. Note the difference between Figure \@ref(fig:figrapportage) and Figure \@ref(fig:figpowerverschil). Figure \@ref(fig:figrapportage) shows the *estimated* proportion based on one sample, while the *real* proportion (the population parameter) is unknown. In Figure \@ref(fig:figpowerverschil) the y-axis corresponds with the (real) proportion in the population and we assume a large number of samples.  

The 2019 report is based on the plots that were measured in the period 2014 - 2018. This means that only a subset of the total sample size was measured at that time, as the total monitoring cycle was 12 years. This subset is still representative for the sample as a whole due to the nature of the sampling design. Only for the more common forest habitat types we already reached the total sample size, as we could make use of the measurements of the Flemish Forest Inventory.

The habitat types are represented by the habitat code and the colour corresponds with the habitat quality evaluation: favourable, unknown or unfavourable. For many habitat types we can already conclude that the quality is unfavourable, even with a sample size much smaller than 170. This is because in many cases the habitat proportion in bad condition is much higher than the 25%-threshold and such large differences can be detected with sufficient power based on smaller sample sizes, as can be seen in Figure \@ref(fig:figpowerverschil) and Table \@ref(tab:tab-powerverschil). Only for two habitat types (2160 and 2190) we need a higher precision in order to conclude if the proportion with bad quality is higher or lower than the 25%-threshold.

In the coming years, it will be more relevant for nature policy to have information on the trend of habitat quality (does the quality of the unfavourable habitat types improve?) than to invest in a higher precision of the status of habitat quality. In other words, it will probably be better to shorten the original monitoring cycle and to start revisiting the measured plots sooner than originally planned. This requires an adjustment of the monitoring design. Therefore we will also perform sample size calculations for trends in habitat quality. This way we can find a good balance between the precision of the habitat quality status and the precision of the trend. 

```{r figrapportage, fig.height = 4.5, fig.width= 5, fig.cap = "Results of the habitat reporting for terrestrial habitat types"}

sim_data0.8 <- sim_data %>%
  filter(power == 0.8) %>%
  filter(n_sim <= 300)

status_types_order <- (status_types %>%
  arrange(n))$type

status_types <- status_types %>%
  mutate(type = factor(type, levels = status_types_order))

ggplot(data = status_types, aes(x = type, y = prop_slecht, ymin = prop_slecht_lci, ymax = prop_slecht_uci, colour = uitspraak)) +
  # geom_line(data = sim_data0.8, aes(x = n_sim, y = p1_low, linetype = factor(alpha))) +
  # geom_line(data = sim_data0.8, aes(x = n_sim, y = p1_high, linetype = factor(alpha))) +
  geom_hline(yintercept = 0.25, linetype = 2,  alpha = 0.5) +
  geom_point( alpha = 0.8, size = 2) +
  geom_errorbar() +
  geom_text(data = status_types, aes(x = type, y = prop_slecht_lci, label = n), alpha = 1, size = 3, vjust = -1.1, colour = "black") +
  labs(x = "Habitat type", y = "Estimated habitat proportion in bad condition", colour = "Conclusion") +
  scale_y_continuous(labels = scales::percent) +
  scale_colour_manual(values = c("Favourable" = inbo_groen, "Unfavourable" = inbo_rood, "Unknown" = inbo_grijs))  +
  theme(axis.text.x = element_text(angle = 90))

```

\needspace{100mm}

## Trends in habitat quality

We will first assume that the habitat condition is a binary variable (good or bad) and analyse which difference between proportions of habitat in bad condition we can detect.

Next, we will evaluate which differences we can detect when a continuous variable is used to express the habitat condition. 

### Difference between proportions

To determine the detectable difference in habitat proportions in bad condition between two points in time, we apply the *McNemar test* [@chow2003] using the function `McNemar.Test` in the R package TrialSize [@zhang2020].

The habitat monitoring scheme consists of permanent plots resulting in paired observations. When the habitat quality in a plot is measured at two time points $t_1$ and $t_2$, we can have one of the following combinations.

```{r}
data.frame(
  t_1 = c("bad", "good"),
  not_good = c("no change", "good -> bad"),
  good = c("bad -> good", "no change")
) %>%
  select("$t_1$" = t_1, "bad" = not_good, good) %>%
  kable(booktabs = TRUE,
        escape = FALSE, 
        format = format_table) %>%
  kable_styling() %>%
 # column_spec(1, bold = TRUE) %>%
  add_header_above(c(" " = 1, "$t_2$" = 2), escape = FALSE)


```

The difference in the habitat proportion in bad condition between $t_1$ and $t_2$ corresponds with the difference in the habitat area that evolves from bad to good ($hab_{01}$) and the area that evolves from good to bad ($hab_{10}$). Figure \@ref(fig:figpowertrend) shows the minimal detectable increase/decrease in the habitat proportion in bad condition as a function of the sample size and the *dynamics* of the habitat type. We define the dynamics as a measure for the amount of habitat for which the condition changes given a total net change in habitat condition:

$$\frac{hab_{01} + hab_{10}}{| hab_{01} - hab_{10}|}$$

Suppose that the net habitat proportion in bad condition decreases with 10 %, then the dynamics are equal to 2 when 15% of the habitat area changes from good to bad and 5% of the habitat area from bad to good. The higher the dynamics, the higher the minimal detectable difference.

```{r figpowertrend, fig.cap = "Minimal detectable effect as a function of sample size and the dynamics with $\\alpha$ = 0.05 and $\\pi$ = 0.80"}

decrease_netto <- seq(0.40, 0.01, by = -0.001)

prop_decrease_bruto <- seq(1.01, 3, by = 0.001)

data_sim <- expand.grid(decrease = decrease_netto,
                        prop_decrease_bruto = prop_decrease_bruto,
                        alpha = c(0.05)) %>%
  mutate(decrease_bruto = decrease * prop_decrease_bruto,
         toename_bruto = decrease_bruto - decrease,
         disc_pairs = decrease_bruto + toename_bruto,
         f = (decrease_bruto + toename_bruto)/ decrease,
         odds = decrease_bruto/toename_bruto,
         alpha = alpha,
         power = 0.8,
         f = disc_pairs/decrease,
         n = McNemar.Test(alpha = alpha,
                          beta = 1 - power,
                          psai =  odds,
                          paid = disc_pairs)) 

data_sim %>%
  mutate(f = round(f, 2)) %>%
  filter(f %in% c(1.02, 1.5, 2)) %>%
  ggplot(aes(x = n, y = decrease, colour = factor(f))) +
  geom_line() +
  scale_x_log10(breaks = c(20, 40, 80 , 170, 400, 1000)) + 
  labs(colour = "Dynamics",
       linetype = expression(alpha),
       y = "Net increase or decrease", 
       x = "Sample size") +
  scale_y_continuous(labels = scales::percent) 


```

Table \@ref(tab:tab-powertrend2) gives an overview of the results for different values for the dynamics. With a sample size of 170 we can detect a difference in the habitat proportion in bad condition of about 5 % when dynamics are low (with $\alpha$ = 0.05 and $\beta$ = 0.8) and a difference of 9 % when dynamics are high.

```{r}
data_sim_tabel_40 <- data_sim %>%
  mutate(f = round(f, 2)) %>%
  filter(f %in% c(1.02, 1.5, 2)) %>%
  group_by(f, alpha) %>%
  slice_min(order_by = abs(n - 40), n = 1) %>%
  ungroup() %>%
  mutate(n_approx = 40)

data_sim_tabel_80 <- data_sim %>%
  mutate(f = round(f, 2)) %>%
  filter(f %in% c(1.02, 1.5, 2)) %>%
  group_by(f, alpha) %>%
  slice_min(order_by = abs(n - 80), n = 1) %>%
  ungroup()%>%
  mutate(n_approx = 80)

data_sim_tabel_170 <- data_sim %>%
  mutate(f = round(f, 2)) %>%
  filter(f %in% c(1.02, 1.5, 2)) %>%
  group_by(f, alpha) %>%
  slice_min(order_by = abs(n - 170), n = 1) %>%
  ungroup()%>%
  mutate(n_approx = 170)

data_sim_tabel_400 <- data_sim %>%
  mutate(f = round(f, 2)) %>%
  filter(f %in% c(1.02, 1.5, 2)) %>%
  group_by(f, alpha) %>%
  slice_min(order_by = abs(n - 400), n = 1) %>%
  ungroup()%>%
  mutate(n_approx = 400)

data_sim_tabel_trend <- bind_rows(data_sim_tabel_40,
                            data_sim_tabel_80,
                            data_sim_tabel_170,
                            data_sim_tabel_400) %>%
  mutate(dynamics = ifelse(f == 1.02, "low dynamics",
                           ifelse(f == 1.5, "medium dynamics",
                                  ifelse(f==2, "high dynamics", NA))),
         dynamics = factor(dynamics, levels = c("low dynamics", "medium dynamics", "high dynamics")),
         decrease = decrease * 100,
         "sample size" = n_approx) %>%
    dplyr::select(decrease, dynamics, "sample size") %>%
  spread(key = "dynamics", value = "decrease")
  

```

```{r tab-powertrend2}
data_sim_tabel_trend %>%
  kable(caption = "Minimal detectable difference as a function of sample size and dynamics (low dynamics = 1.02, medium dynamics =  1.5 and high dynamics = 2) with $\\alpha$ = 0.05 and $\\pi$ = 0.80",
        booktabs = TRUE,
        align = "c", 
        format = format_table) %>%
  kable_styling() %>%
  add_header_above(c(" " = 1, "net decrease or increase (%)" = 3))
  
```

It is hard to predict which values for the dynamics can be expected for the different habitat types. However, data from the Flemish Forest Inventory might give us an idea for the forest habitat types. Table \@ref(tab:dynamicsbostypes) shows the number of paired observations and the number of plots that evolve from bad to good quality ($n_{01}$) and from good to bad quality ($n_{10}$). For habitat type 9120 the dynamics are equal to 2.2. For the other forest habitat types we can not get a reliable estimate of the dynamics as the sample size is too low or there is no net difference. Anyway, assuming high dynamics in sample size calculation does not seem exaggerated.

```{r}

data_bos <- read.csv2("data/StatusHabitatvlek_bos_trend.csv") %>%
  filter(Versie == "Versie 3") %>%
  select(id = ID, periode = Periode, type = Habitattype, sac = SBZH, stratum_weigh = StratumWeight, status = Status_habitatvlek, aandeel_gunstig = AandeelGunstig, id_group = IDGroup) %>%
  mutate(plot_id = str_sub(id, end = -3)) %>%
  mutate(periode = str_c("periode", periode)) %>%
  filter(type != 9110) %>%
  mutate(type = as.character(type),
         type = ifelse(type == "91", "91E0", as.character(type))) %>%
  group_by(id_group) %>%
  mutate(paired = n() > 1) %>%
  ungroup()

check <- data_bos %>%
  group_by(periode, type) %>%
  summarise(n_gepaard = sum(paired) )

```

```{r}
dynamics_forest <- data_bos %>%
 # filter(paired) %>%
  dplyr::select(type, plot_id, status, periode) %>%
  spread(key = periode, value = status) %>%
  group_by(type) %>%
  summarise(n_gepaard = n_distinct(plot_id),
            n_00 = sum(periode1 == 0 & periode2 == 0),
         n_11 = sum(periode1 == 1 & periode2 == 1),
         n_01 = sum(periode1 == 0 & periode2 == 1),
         n_10 = sum(periode1 == 1 & periode2 == 0),
         ) %>%
  ungroup() %>%
  mutate(dynamics = ifelse(abs(n_01 - n_10) > 0, (n_01 + n_10)/abs(n_01 - n_10), NA),
         verschil_slecht = (n_10 - n_01)/n_gepaard * 100) 
```

```{r dynamicsbostypes}
options(knitr.kable.NA = '')

dynamics_forest %>%
  select("habitat type" = type, "$n_{paired}$" = n_gepaard, "$n_{01}$" = n_01 , "$n_{10}$" = n_10, dynamics) %>%
  kable(caption = "Observed dynamics for forest habitat types based on data from the Flemish Forest Inventory",
        booktabs = TRUE,
        escape = FALSE, 
        format = format_table) %>%
  kable_styling()
```

### Differences between habitat quality indicators

Continuous, discrete or categorical variables that measure habitat quality are more sensitive to changes than a binary variable (good or bad habitat condition). Examples of such variables are:

-   values of habitat quality indicators as listed in [@oosterlynck2020]
-   the proportion of habitat quality indicators for a certain habitat type that are favourable
-   a continuous habitat quality index based on the distance to the threshold values for the different habitat quality indicators of a habitat type

```{r}
sim_data <- expand.grid(n = 20:400,
                        sd_difference = c(5, 10, 15),
                        alpha = c(0.050)) %>%
  group_by(n, sd_difference, alpha) %>%
  mutate(d = (pwr.t.test(n = n,
                        power = 0.80,
                        sig.level = alpha,
                        type = "paired",
                        alternative = "two.sided"
                        ))$d) %>%
  ungroup() %>%
  mutate(mean_difference = d * sd_difference)
```

For the exploration of the detectable effect size as a function of the sample size, we assume continuous variables with a normal distribution. This enables a straightforward power analysis based on the paired t-test using the `pwr.t.test` of the pwr package [@champely2020]. Although not all habitat quality indicators will have normally distributed paired differences, we believe that this method will give us a general idea on the relationship between detectable effect size and sample size.

We will verify some of the results of the paired t-test, by applying a more advanced method to calculate the power using the R package `simr` [@green2016]. The advantage of this method is that we can make use of other distributions, such as the poisson distribution and the binomial distribution. More details are provided in Appendix \@ref(annexsimr)  .    

#### Sample size calculation for continuous variables based on the paired t-test

When applying the paired t-test, the detectable effect size ($d$) equals the mean of the differences between paired observations divided by the standard deviation of the differences. Figure \@ref(fig:fig-verschil-cont) and Table \@ref(tab:powerd) show the detectable effect size as a function of the sample size.

```{r fig-verschil-cont, fig.cap = "Detectable effect size ($d$) as a function of the sample size with $\\pi = 0.8$"}
sim_data %>%
  ggplot(aes(x = n, y = d,  )) +
  geom_line() +
  labs(x = "Sample size",
       y = "Detectable effect") 
```

```{r powerd}
sim_data %>%
  filter(n %in% c(40, 80, 170, 400)) %>%
  mutate(d = round(d, 3)) %>%
  arrange(alpha, n) %>%
  dplyr::select("Sample size" = n, "Detectable effect" = d) %>%
  unique() %>%
  kable(caption = "Detectable effect as a function of sample size and significance level with $\\pi=0.80$",
        booktabs = TRUE, 
        format = format_table) %>%
  kable_styling(full_width = FALSE) 
  
```

The detectable effect is a quite abstract measure and  not so easy to interpret. To make things more  comprehensible, in Figure \@ref(fig:fig-verschil-cont2) and Table \@ref(tab:tab-powertrend) we show a variable expressed as a percentage with values between 0 and 100. This could be for example the proportion of habitat indicators that are favourable on a certain location or the vegetation cover of a certain species group. With a sample size of 170 we can detect a mean difference of approximately 1 to 3 % depending on the standard deviation of the difference. With a sample size of 80 this is approximately 1.5 to 5 %.

```{r fig-verschil-cont2, fig.cap = "Detectable mean difference as a function of sample size and standard deviation of the difference for a variable that is expressed as a percentage with $\\alpha=0.05$ and $\\pi=0.80$"}
sim_data %>%
  filter(alpha == 0.05) %>%
ggplot(aes(x = n, y = mean_difference/100, colour = factor(sd_difference))) +
  geom_line() +
  scale_x_log10(breaks = c(20, 40, 80 , 170, 400, 1000)) + 
  scale_y_continuous(labels = scales::percent) +
  labs(linetype = expression(alpha),
       colour = "Standaard deviation \nof difference",
       y = "Mean difference",
       x = "Sample size") 
```

```{r tab-powertrend}
sim_data %>%
  filter(n %in% c(40, 80, 170, 400)) %>%
  mutate(mean_difference = round(mean_difference, 1)) %>%
  dplyr::select(n, alpha, sd_difference, mean_difference) %>%
  filter(alpha == 0.05) %>%
  spread(key = "sd_difference", value = "mean_difference") %>%
  arrange(alpha, n) %>%
  dplyr::select("sample size" = n, everything()) %>%
  select(-alpha) %>%
  kable(caption = "Detectable mean difference as a function of sample size and standard deviation of the difference with $\\alpha=0.05$ and $\\pi=0.80$",
        booktabs = TRUE, 
        format = format_table) %>%
  kable_styling() %>%
  add_header_above(c(" " = 1, "sd difference" = 3))
  
```

We will further explore the data of the Flemish Forest Inventory [@wouters_ontwerp_2008] to get a rough idea of the standard deviation of differences we can expect for habitat quality indicators. 

#### Example 1: trend in proportion of favourable indicators for forest habitat types

The Forest Inventory data allows the calculation and evaluation of most habitat quality indicators defined in [@oosterlynck2020] for plots that are located in a forest habitat types. Based on this, we can determine the proportion of the total number of indicators that are favourable per plot. Figure \@ref(fig:figbos) shows this proportion for plots that were measured in both the first period (1997-1998) and the second period (2009-2018) of the Flemish Forest Inventory.

```{r figbos, fig.height = 4.2, fig.width= 5.3, fig.cap= "Proportion of habitat quality indicators that are favourable per plot, for both periods of the Flemish Forest Inventory"}
data_bos %>%
  filter(paired) %>%
  dplyr::select(plot_id, periode, type, aandeel_gunstig) %>%
  mutate(aandeel_gunstig = aandeel_gunstig/100) %>%
  spread(key = "periode", value = "aandeel_gunstig") %>%
  ggplot(aes(x= periode1, y = periode2)) +
  #geom_point(alpha = 0.6) +
  stat_sum(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  labs(x = "Period 1", y = "Period 2") +
  scale_x_continuous(labels = scales::percent, limits = c(0,1)) +
  scale_y_continuous(labels = scales::percent, limits = c(0,1)) +
  scale_size("Number \nof plots") +
  facet_wrap(~type)
```
```{r eval = FALSE}
data_bos %>%
  filter(paired) %>%
  dplyr::select(plot_id, periode, type, aandeel_gunstig) %>%
  spread(key = periode, value = aandeel_gunstig) %>%
  mutate(verschil_aandeel_gunstig = periode2 - periode1) %>%
  ggplot(aes(x = verschil_aandeel_gunstig)) +
  geom_bar() +
  facet_wrap(~type)
```



```{r}
verschil_aandeel_gunstig <- data_bos %>%
  filter(paired) %>%
  dplyr::select(plot_id, periode, type, aandeel_gunstig) %>%
  spread(key = periode, value = aandeel_gunstig) %>%
  mutate(verschil_aandeel_gunstig = periode2 - periode1) %>%
  group_by(type) %>%
  summarise(n = n(),
            mean_diff = round(mean(verschil_aandeel_gunstig), 2),
            sd_diff = round(sd(verschil_aandeel_gunstig), 2),
            d = round(mean_diff/sd_diff, 3),
            t.test_p_value = round((t.test(Pair(periode1, periode2) ~ 1))$p.value, 5),
            uitspraak = ifelse(t.test_p_value > 0.05, "unknown",
                               ifelse(mean_diff < 0, "decrease", "increase"))
           ) %>%
  ungroup() 

verschil_aandeel_gunstig_tot <- data_bos %>%
  filter(paired) %>%
  dplyr::select(plot_id, periode, type, aandeel_gunstig) %>%
  spread(key = periode, value = aandeel_gunstig) %>%
  mutate(verschil_aandeel_gunstig = periode2 - periode1) %>%
  mutate(type = "91xx") %>%
  group_by(type) %>%
  summarise(n = n(),
            mean_diff = round(mean(verschil_aandeel_gunstig), 2),
            sd_diff = round(sd(verschil_aandeel_gunstig), 2),
            d = round(mean_diff/sd_diff, 3),
            t.test_p_value = round((t.test(Pair(periode1, periode2) ~ 1))$p.value, 5),
            uitspraak = ifelse(t.test_p_value > 0.05, "unknown",
                               ifelse(mean_diff < 0, "decrease", "increase"))
           ) %>%
  ungroup() 

verschil_aandeel_gunstig <- verschil_aandeel_gunstig %>%
  bind_rows(verschil_aandeel_gunstig_tot)
```

Table \@ref(tab:tabmeandif) gives the mean difference of the proportion of favourable indicators between period 2 and period 1, the standard deviation of the difference and the detectable effect size ($d$). The table also shows the p-value of the paired t-test and the conclusion that can be drawn per type and for all types together ('91xx'). The standard deviation of the differences ranges approximately from 7% to 11%.


```{r tabmeandif}
verschil_aandeel_gunstig %>%
  dplyr::select("habitat type" = type, "sample size" = n, "mean difference" = mean_diff, "sd difference" = sd_diff, d, "p-value" = t.test_p_value, conclusion = uitspraak) %>%
  kable(caption = "Observed difference of the proportion of favourable indicators for forest habitat types", 
        booktabs = TRUE, 
        format = format_table,
        digits = 3
        ) %>%
  kable_styling() %>%
  column_spec(4, width = "2.5cm")

```
When the estimated effect size ($d$) would correspond with the true effect size, a sample size of 80 would be sufficient to detect the effect for 9130 and 9160 as can be seen in Figure \@ref(fig:fig-verschil-cont) and Table \@ref(tab:powerd). For the other habitat types we would need a higher sample size, but these are quite small effect sizes. 

Table \@ref(tab:tabdiffdetect1) shows the range of the minimum detectable difference in proportion of favourable indicators that can be expected for different sample sizes with a power $\pi$ of 0.80. This is based on the range of values for the standard deviation of differences as shown in Table \@ref(tab:tabmeandif).    

```{r diffdetect1}

verschil_aandeel_gunstig_detecteerbaar <- sim_data %>%
  filter(n %in% c(40, 80, 170)) %>%
  mutate(d = round(d, 2)) %>%
  filter(alpha == 0.05) %>%
  distinct(n, d, alpha) %>%
  mutate(sd_diff_low = floor(min(verschil_aandeel_gunstig$sd_diff)),
         sd_diff_high = ceiling(max(verschil_aandeel_gunstig$sd_diff)),
         mean_diff_low = round(d * sd_diff_low, 1),
         mean_diff_high = round(d * sd_diff_high, 1),
         sd_diff_range = str_c(sd_diff_low, " - ", sd_diff_high),
         mean_diff_range = str_c(mean_diff_low, "% - ", mean_diff_high, "%")) 

verschil_aandeel_gunstig_9210 <- verschil_aandeel_gunstig %>%
  filter(type == "9120")

verschil_aandeel_gunstig_9210_detecteerbaar <- sim_data %>%
  filter(n %in% c(40, 80, 170)) %>%
  mutate(d = round(d, 2)) %>%
  filter(alpha == 0.05) %>%
  distinct(n, d, alpha) %>%
  mutate(sd_diff = verschil_aandeel_gunstig_9210$sd_diff,
         mean_diff = d * sd_diff) 

```

```{r tabdiffdetect1}
verschil_aandeel_gunstig_detecteerbaar %>%
  select( "sd difference" = sd_diff_range, "sample size" = n, d, "mdd" = mean_diff_range) %>%
  kable(caption = "Minimum detectable differencev(mdd) in proportion of favourable indicators that can be expected for different sample sizes with a power $pi$ of 0.80",
        booktabs = TRUE, 
        format = format_table) %>%
  kable_styling() %>%
  collapse_rows(1)
  
```

```{r, cache = TRUE}

calc_power_simr <- function(model, sim_n, sim_effect) {
  
  fixef(model)["periodeperiode2"] <- sim_effect
  model_sim <- extend(model, along = "id_group", n = sim_n)
  sim <- powerSim(model_sim, nsim = 250)
  
  summary <- summary(sim)
  
  result_temp <- data.frame(n = sim_n,
                            effect = sim_effect,
                            power = summary$mean,
                            lcl = summary$lower,
                            ucl = summary$upper)
  
  return(result_temp)
  
}

calc_power_simr_df <- function(model, df) {
  
  results <- NULL
  
  for(i in 1:nrow(df)) {
    
    result_temp <- calc_power_simr(model, df[i,]$n, df[i,]$effect)
    
    results <- bind_rows(results,
                         result_temp)
    
  }
  
  return(results)
  
}


data_bos_paired <- data_bos %>%
  filter(paired) %>%
  filter(type == "9120") %>%
  mutate(aandeel_gunstig = aandeel_gunstig/100)

model_9120 <- lmer(formula = aandeel_gunstig ~ periode + (1|id_group),
              data = data_bos_paired)

```

```{r, eval= FALSE}

verschil_aandeel_gunstig_9210_detecteerbaar <- verschil_aandeel_gunstig_9210_detecteerbaar %>%
  mutate(effect = mean_diff/100)

set.seed(675)

verschil_aandeel_gunstig_9210_simr <- calc_power_simr_df(model_9120, verschil_aandeel_gunstig_9210_detecteerbaar) 

write_csv2(verschil_aandeel_gunstig_9210_simr, "../output/verschil_aandeel_gunstig_9210_simr.csv")
```


```{r, cache = FALSE}
verschil_aandeel_gunstig_9210_simr <- read_csv2("../output/verschil_aandeel_gunstig_9210_simr.csv")

compare_power_aandeel_gunstig_9210 <- verschil_aandeel_gunstig_9210_simr %>%
  bind_cols(select(verschil_aandeel_gunstig_9210_detecteerbaar, mean_diff)) %>%
  select(n, effect_size = mean_diff, power_simr = power, lcl, ucl)
```

To verify the results we calculated the power for habitat type 9120 using the `simr` package. The power is close to 0.8 for all combinations of sample size and effect size, confirming the results in Table \@ref(tab:tabdiffdetect1). The estimated power using `simr` is shown in Table \@ref(tab:tabcompsimr1) in Appendix \@ref(annexsimr).     

\needspace{50mm}

#### Example 2: trend in indicator values for forest habitat types

Here we use the Forest Inventory data to get a rough idea on the detectable effect size for individual habitat quality indicators. We analyse four indicators:

- number of key species in the herb layer
- proportion of key species in the herb layer
- total cover of encroachment species
- total cover of invasive alien species in the shrub and tree layer


```{r}
voorwaarden_bos <- read.csv2("data/voorwaarden_bos.csv", stringsAsFactors = FALSE) 

data_bos_paired <- data_bos %>%
  filter(paired)

voorwaarden_bos_select <- voorwaarden_bos %>%
  mutate(plot_id = str_c(Meetnet, IDPlots),
         periode = str_c("periode", Periode),
         type = ifelse(Habitattype == 91, "91E0", as.character(Habitattype))) %>%
  filter(Meetnet == "VBI") %>%
  filter(Voorwaarde %in% c("aandeel sleutelsoorten kruidlaag", "aantal sleutelsoorten kruidlaag", "bedekking verruiging totaal", "bedekking invasieve exoten boom- en struiklaag")) %>%
  mutate(voorwaarde_engels = ifelse(Voorwaarde == "aandeel sleutelsoorten kruidlaag", "proportion of key species in the herb layer",
                                    ifelse(Voorwaarde == "aantal sleutelsoorten kruidlaag", "number of key species in the herb layer",
                                           ifelse(Voorwaarde == "bedekking verruiging totaal", "total cover of encroachment species",
                                                  "total cover of invasive alien species in the shrub and tree layer")))) %>%
  filter(Versie == "Versie 3") %>%
  filter(plot_id %in% data_bos_paired$plot_id) %>%
  dplyr::select(plot_id, periode, type, voorwaarde = Voorwaarde, voorwaarde_engels, waarde = Waarde) %>%
  mutate(waarde = str_replace(waarde, ",", "."),
         waarde = as.numeric(waarde),
         waarde = ifelse(voorwaarde == "aandeel sleutelsoorten kruidlaag", pmin(100, waarde), waarde)) %>%
  spread(key = "periode", value = "waarde") %>%
  mutate(verschil_voorwaarde = periode2 - periode1)

# na te kijken: aandeel sleutelsoorten kruidlaag > 100?
```

Figure \@ref(fig:indicatorenbos) shows the indicator values for both periods of the Flemish Forest Inventory.

```{r indicatorenbos, fig.width = 7.5, fig.height= 6, fig.cap = "Indicator values for both periods of the Flemish Forest Inventory "}

voorwaarden_bos_select %>%
  mutate(voorwaarde_engels = str_wrap(voorwaarde_engels)) %>%
  ggplot(aes(periode1, periode2, colour = type)) +
  geom_point(alpha = 0.5) +
  facet_wrap( ~ voorwaarde_engels, scales = "free", ncol = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  labs(x = "Period 1", y =" Period 2")

```

```{r indicatorenbos2,  fig.height= 7, fig.cap = "Indicatorwaarden voor de meetpunten van de Vlaamse bosinventarisatie", eval=FALSE}

voorwaarden_bos_select %>%
  ggplot(aes(x = verschil_voorwaarde)) +
  geom_histogram(alpha = 0.5) +
  facet_wrap( ~ voorwaarde_engels, scales = "free", ncol = 2)
```

```{r}
verschil_voorwaarden <- voorwaarden_bos_select %>%
  group_by(type, voorwaarde_engels) %>%
  summarise(n = n(),
            mean_diff = round(mean(verschil_voorwaarde), 2),
            sd_diff = round(sd(verschil_voorwaarde), 2),
            d = round(mean_diff/sd_diff, 3),
            t.test_p_value = round((t.test(Pair(periode1, periode2) ~ 1))$p.value, 5),
             uitspraak = ifelse(t.test_p_value > 0.05, "unknown",
                               ifelse(mean_diff < 0, "decrease", "increase"))
           ) %>%
  ungroup() 

verschil_voorwaarden_tot <- voorwaarden_bos_select %>%
  mutate(type = "91xx") %>%
  group_by(type, voorwaarde_engels) %>%
  summarise(n = n(),
            mean_diff = round(mean(verschil_voorwaarde), 2),
            sd_diff = round(sd(verschil_voorwaarde), 2),
            d = round(mean_diff/sd_diff, 3),
            t.test_p_value = round((t.test(Pair(periode1, periode2) ~ 1))$p.value, 5),
             uitspraak = ifelse(t.test_p_value > 0.05, "unknown",
                               ifelse(mean_diff < 0, "decrease", "increase"))
           ) %>%
  ungroup() 

verschil_voorwaarden <- verschil_voorwaarden %>%
  bind_rows(verschil_voorwaarden_tot)
```

\needspace{150mm}

Table \@ref(tab:tabindicatoren) gives the observed effect size $d$ for the habitat quality indicators. The table also shows the p-value of the paired t-test and the conclusion that can be drawn per type and for all types together ('91xx'). We can also compare $d$ with the values in Table \@ref(tab:powerd) to evaluate which differences can be detected for different sample sizes (assuming that the observed $d$ values are the true values). 

```{r tabindicatoren}
verschil_voorwaarden %>%
  filter(type %in% c("9120", "91E0", "91xx")) %>%
  dplyr::select("habitat type" = type, "sample size" = n, indicator = voorwaarde_engels, "mean diff" = mean_diff, "sd diff" = sd_diff, d, "p-value" = t.test_p_value, conclusion = uitspraak) %>%
  kable(caption = "Observed effect size ($d$) for four habitat quality indicators",
        booktabs = TRUE, 
        format = format_table,
        digits = 3) %>%
  kable_styling() %>%
  column_spec(3, width = "4cm") %>%
  column_spec(c(4,5,7), width = "1cm") %>%
  column_spec(c(1, 2), width = "0.8cm") %>%
  collapse_rows(c(1,2))
```


```{r tabdiffdetect2}

verschil_voorwaarden_range <- verschil_voorwaarden %>%
  filter(type %in% c("91E0", "9120", "91xx")) %>%
  group_by(voorwaarde_engels) %>%
  summarise(sd_diff_low = floor(min(sd_diff)),
         sd_diff_high = ceiling(max(sd_diff))) %>%
  ungroup() %>%
  tidyr::expand(nesting(voorwaarde_engels, sd_diff_low, sd_diff_high), n = c(40, 80, 170))
  
verschil_voorwaarden_detecteerbaar <- sim_data %>%
  filter(n %in% c(40, 80, 170)) %>%
  mutate(d = round(d, 2)) %>%
  filter(alpha == 0.05) %>%
  distinct(n, d) %>%
  left_join(verschil_voorwaarden_range, by = "n") %>%
  mutate(mean_diff_low = round(d * sd_diff_low, 1),
         mean_diff_high = round(d * sd_diff_high, 1),
         sd_diff_range = str_c(sd_diff_low, " - ", sd_diff_high),
         mean_diff_range = str_c(mean_diff_low, " - ", mean_diff_high)) 

```


```{r tabdiffdetect3}

voorwaarde_9120_mean_period <- voorwaarden_bos_select %>%
  filter(type == "9120") %>%
  group_by(voorwaarde_engels) %>%
  summarise(mean_p1 = mean(periode1),
            mean_p2 = mean(periode2)) %>%
  ungroup()

verschil_voorwaarden_9120 <- verschil_voorwaarden %>%
  filter(type == "9120") %>%
  tidyr::expand(nesting(voorwaarde_engels, sd_diff), n = c(40, 80, 170))
  
verschil_voorwaarden_detecteerbaar_9120 <- sim_data %>%
  filter(n %in% c(40, 80, 170)) %>%
  mutate(d = round(d, 2)) %>%
  filter(alpha == 0.05) %>%
  distinct(n, d) %>%
  left_join(verschil_voorwaarden_9120, by = "n") %>%
  left_join(voorwaarde_9120_mean_period, by = "voorwaarde_engels") %>%
  mutate(mean_diff_detect = d * sd_diff,
         log_effect = log(mean_diff_detect + mean_p1) - log(mean_p1)) 

```

\needspace{200mm}

Table \@ref(tab:tabmind) shows the range of  the detectable mean difference for the habitat quality indicators that can be expected for different sample sizes. It is based on the range of observed values for the standard deviation of the indicators as shown in Table \@ref(tab:tabindicatoren).     


```{r tabmind}
verschil_voorwaarden_detecteerbaar %>%
  arrange(voorwaarde_engels) %>%
  mutate(voorwaarde_engels = str_replace(voorwaarde_engels, "proportion", "proportion (%)"),
         voorwaarde_engels = str_replace(voorwaarde_engels, "cover", "cover (%)")) %>%
  select(indicator = voorwaarde_engels, "sd difference" = sd_diff_range, "sample size" = n,  mdd = mean_diff_range) %>%
  kable(caption = "Expected range for minimum detectable difference (mdd) for habitat quality indicators",
        booktabs = TRUE, 
        format = format_table) %>%
  kable_styling() %>%
  column_spec(1, width = "4cm") %>%
  column_spec(c(2,3,4), width = "2cm") %>%
  collapse_rows(c(1, 2))
```

\needspace{70mm}

We used the `simr` package to verify the results in Table \@ref(tab:tabmind). The estimated power based on `simr` was higher than 0.80 for the number of key species and similar to 0.80 for the total cover of encroachment species for habitat type 9120 (Table \@ref(tab:tabcompsimr2) and Table \@ref(tab:tabcompsimr3). This indicates that the minimal detectable differences in \@ref(tab:tabmind) are valid or even slightly smaller in case of the number of key species.




```{r}
n_keyspecies_9120 <- voorwaarden_bos_select %>%
  filter(type == "9120") %>%
  filter(voorwaarde == "aantal sleutelsoorten kruidlaag") %>%
  select(-verschil_voorwaarde) %>%
  pivot_longer(cols = c("periode1", "periode2"), names_to = "periode", values_to = "n") %>%
  mutate(periode_n = str_remove(periode, "periode"),
         periode = factor(periode)) %>%
  mutate(id = str_c(plot_id, "_", periode_n)) %>%
  left_join(select(data_bos_paired, id, id_group), by = "id")

model_ks_9120_poisson <- glmer(formula = n ~ periode + (1|id_group),
              data = n_keyspecies_9120,
              family = "poisson")

verschil_keyspecies_detecteerbaar_9120 <- verschil_voorwaarden_detecteerbaar_9120 %>%
  filter(voorwaarde_engels == "number of key species in the herb layer") %>%
  mutate(effect = log_effect)

data_simr <- tibble(n = c(40, 80, 170)) %>%
  tidyr::expand( n, nspecies_p1 = c(2, 3, 5), increase = seq(0.2, 0.5, by = 0.05)) %>%
  mutate(nspecies_p2 = nspecies_p1 + increase*nspecies_p1,
         effect = log(nspecies_p2) - log(nspecies_p1))
```

```{r, eval = FALSE}
verschil_keyspecies_simr <- calc_power_simr_df(model_ks_9120_poisson, verschil_keyspecies_detecteerbaar_9120)

write_csv2(verschil_keyspecies_simr, "../output/verschil_keyspecies_simr")

simr_result <- calc_power_simr_df(model_ks_9120_poisson, df = data_simr)

simr_result2 <- simr_result %>%
  bind_cols(select(data_simr, nspecies_p1, nspecies_p2, increase))

write_csv2(simr_result2, "../output/simulation_nspecies.csv")
```

```{r}
verschil_keyspecies_simr <- read_csv2("../output/verschil_keyspecies_simr")

compare_power_keyspecies_9210 <- verschil_keyspecies_simr %>%
  bind_cols(select(verschil_keyspecies_detecteerbaar_9120, mean_diff_detect)) %>%
  select(n, effect_size = mean_diff_detect, power_simr = power, lcl, ucl)
```

```{r, eval = FALSE}

simr_result2 <- read_csv2("../output/simulation_nspecies.csv")

simr_result2 %>%
  ggplot(aes(y = power, ymin = lcl, ymax = ucl, x = factor(n), group = factor(increase), colour = factor(increase))) +
  geom_point() +
  geom_errorbar(width = 0.1, alpha = 0.8) +
  geom_line() +
  geom_hline(yintercept = 0.8, linetype = 2) +
  facet_wrap(~nspecies_p1) +
  labs(colour = "Relative increase", x = "Sample size")
```




```{r}

encroachement_9120 <- voorwaarden_bos_select %>%
  filter(type == "9120") %>%
  filter(voorwaarde == "bedekking verruiging totaal") %>%
  select(-verschil_voorwaarde) %>%
  pivot_longer(cols = c("periode1", "periode2"), names_to = "periode", values_to = "cover") %>%
  mutate(periode_n = str_remove(periode, "periode"),
         periode = factor(periode),
         cover_adj = ifelse(cover == 0, 0.001, cover/100),
         cover_scale = qlogis(cover_adj)) %>%
  mutate(id = str_c(plot_id, "_", periode_n)) %>%
  left_join(select(data_bos_paired, id, id_group), by = "id")

model_encroachement_9120 <- lmer(formula = cover_scale ~ periode + (1|id_group),
              data = encroachement_9120)

model_encroachement_9120_summary <- summary(model_encroachement_9120)


verschil_encroachment_detecteerbaar_9120 <- verschil_voorwaarden_detecteerbaar_9120 %>%
  filter(voorwaarde_engels == "total cover of encroachment species") %>%
  mutate (effect = qlogis(mean_p1/100 + mean_diff_detect/100) - qlogis(mean_p1/100))


data_simr <- tibble(n = c(40, 80, 170)) %>%
  tidyr::expand( n, nspecies_p1 = c(2, 3, 5), increase = seq(0.1, 0.4, by = 0.05)) %>%
  mutate(nspecies_p2 = nspecies_p1 + increase*nspecies_p1,
         effect = log(nspecies_p2) - log(nspecies_p1))

```


```{r, eval = FALSE}

verschil_encroachment_simr <- calc_power_simr_df(model_encroachement_9120, verschil_encroachment_detecteerbaar_9120)

write_csv2(verschil_encroachment_simr, "../output/verschil_encroachment_simr.csv")
```

```{r}

verschil_encroachment_simr <- read_csv2("../output/verschil_encroachment_simr.csv")

compare_power_encroachment_9210 <- verschil_encroachment_simr %>%
  bind_cols(select(verschil_encroachment_detecteerbaar_9120, mean_diff_detect)) %>%
  select(n, effect_size = mean_diff_detect, power_simr = power, lcl, ucl)
```


