
# Sample design for terrestrial habitat types

## Habitat types and subtypes included in MHQ 

As mentioned in @westra2014, MHQ does not contain a monitoring scheme for all habitat types or habitat subtypes that occur in Flanders. For several habitat types a habitat quality monitoring scheme is not considered as a feasible or appropriate approach:

+ very rare terrestrial habitat types (defined as habitat types with an area lower than 10 ha or occurring on less than 15 locations): a monitoring scheme would result in a very high density of plots which might impact the vulnerable vegetation;
+ estuaries (type 1130): the habitat quality has to be evaluated on a landscape scale;
+ mudflats (type 1140): this habitat type is not included in @oosterlynck2020;
+ very dynamic habitat types (type 1130, 2110 and 7150): a basic habitat quality assessment combined with habitat area mapping is considered a more appropriate approach;
+ habitat type 6430: a very patchy habitat type for which an accurate sample frame is not available;
+ habitat type 9110: habitat type only occurring in the continental biogeographical region of Flanders and it covering only a small area compared to the total area of 9110 in Belgium. Therefore a separate estimate for Flanders is not considered very relevant.

An oversampling is considered for all habitat subtypes except for the very rare terrestrial subtypes (defined as habitat subtypes with an area lower than 10 ha or occurring on less than 15 locations). As for dynamic habitat types, we intend to perform a basic habitat quality assessment for very rare terrestrial habitat types and subtypes is combination with the habitat area mapping.

Appendix \@ref(annex2) shows for which habitat types a monitoring scheme was developed and for which subtypes oversampling is considered. The area and the number of locations (to determine the very rare habitat types and subtypes) is derived from the standardized map of habitat types in Flanders [@westra2021]. This data source is a processed version of the Natura 2000 habitat map [@inbo2020]. The standardized map is processed in a way it can be easily imported and analysed in R. It can be imported in R by using the function `read_habitatmap_stdized()` of the n2khab package [@vanderhaeghe2022].

In the original design of @westra2014 a monitoring scheme was also planned for habitat type 5130 and 6430. However these monitoring schemes were not implemented. Habitat type 5130 is considered a very rare habitat type (see Appendix \@ref(annex2)) and for habitat type 6430 a monitoring scheme is not feasible as mentioned above.    

check 2190_mp




```{r}
habmap_terr <- read_habitatmap_terr()

habmap_types <- habmap_terr$habitatmap_terr_types %>%
  filter(phab > 10) %>%
  filter(str_sub(type, 1,3) != "rbb")

habmap_polygons <- habmap_terr$habitatmap_terr_polygons

types <- read_types() %>%
  select(type, typeclass, typelevel, main_type)

exclude_type <- c("9110", "1130", "1140", "1310", "3270", "5130", "2110", "6120", "6430", "7150")

 n_polygons_min = 15
 area_min = 10

sac <- read_admin_areas(dsn = "sac") %>%
  group_by(sac_code) %>%
  summarise(n_pol = n()) %>%
  ungroup()

habmap_polygons <- habmap_polygons %>%
  st_join(sac) %>%
  mutate(sac = !is.na(sac_code)) %>%
  mutate(area_polygon = drop_units(st_area(geom))) %>%
  st_drop_geometry()

types_area_strata <- habmap_types %>%
  left_join(habmap_polygons, by = "polygon_id") %>%
  left_join(types, by = "type") %>%
  group_by(type, typelevel, typeclass, main_type, sac) %>%
  summarise(area_subtype_sac = sum(phab/100 * area_polygon)/10000,
            area_polygons_subtype_sac = sum(area_polygon)/10000,
            n_polygons_subtype_sac = n_distinct(polygon_id),
            prop_subtype_sac = area_subtype_sac/area_polygons_subtype_sac) %>%
  ungroup() %>%
  group_by(type) %>%
  mutate(area_subtype = sum(area_subtype_sac),
            area_polygons_subtype = sum(area_polygons_subtype_sac),
            n_polygons_subtype = sum(n_polygons_subtype_sac),
            prop_subtype = area_subtype/area_polygons_subtype) %>%
  ungroup() %>%
  group_by(main_type, sac) %>%
  mutate(area_type_sac = sum(area_subtype_sac),
            area_polygons_sac = sum(area_polygons_subtype_sac),
            n_polygons_sac = sum(n_polygons_subtype_sac),
            prop_type_sac = area_type_sac/area_polygons_sac) %>%
  ungroup() %>%
  group_by(main_type) %>%
  mutate(area_type = sum(area_subtype_sac),
            area_polygons = sum(area_polygons_subtype_sac),
            n_polygons = sum(n_polygons_subtype_sac),
            prop_type = area_type/area_polygons,
            area_type_insac = sum(area_subtype_sac * (sac))
          ) %>%
  ungroup() %>%
  mutate(prop_type_insac = area_type_insac/area_type) %>%
  mutate(sample_type =  ifelse(main_type %in% exclude_type, FALSE,
                               ifelse(typeclass == "FW", TRUE, (n_polygons > n_polygons_min) & (area_type > area_min))),
         extra_subtype = ifelse((main_type %in% exclude_type) | (typelevel != "subtype"), FALSE,
                               ifelse(typeclass == "FW", TRUE, (n_polygons_subtype > n_polygons_min) & (area_subtype > area_min)))
         )

```



## Scenario's for sample design {#scenarioterr}

We will calculate the total number of sampling units that have to be measured per year for different scenario's. The scenarios are characterised by following parameters:

+ monitoring cycle: the number of years in which all sampling units have to measured
+ $n_{fl}$: the sample size per habitat type for the whole of Flanders
+ $n_{sac}$: the sample size per habitat type within the Special Areas of Conservation (SAC)
+ $n_{subtype}$: the (minimum) sample size per habitat subtype for the whole of Flanders 

The method for the calculation of the total sample size per habitat type is explained in @westra2014. In a first step a finite population corrections is applied on $n_{fl}$, $n_{sac}$ and $n_{subtype}$. This is because the calculated sample sizes (chapter ) are valid for infinite populations. Yet, most habitat types or subtypes cover a relatively small area and can therefore be considered as a finite population. The finite sample size $n_{finite}$ is calculated as follows:

$$n_{finite}=\frac{Nn_{infinite}}{n_{infinite} + (N-1)}$$
with 

$$n_{infinite} = \text{infinite sample size}$$
$$N = \text{total population size}$$
For the terrestrial habitat types a population unit corresponds with a 32m x 32m grid cell and the total population size is calculated by dividing the total area of the population by the area of a grid cell (approximately 1 ha). Next we distribute $n_{fl_{finite}}$ over the strata *inside sac* and *outside sac* proportional to the area of both strata, and we calculate the extra number of sampling units to reach $n_{sac_{finite}}$ inside sac. Then the total number of sampling units inside and outside sac are distributed over the subtypes proportional to their area to obtain the expected sample size for each stratum. The required sample size per subtype $n_{subtype_{finite}}$ is also distributed over the strata inside sac and outside sac proportional to the area. Finally the extra number of sampling units to reach the required sample size per stratum is calculated.


```{r}


calc_finitsamplesize_terrhab <- function(n_type_flanders = 80, n_type_sac = 170, n_subtype_flanders = 80, types_area_strata = types_area_strata) {
  
  types_samplesize_detailed <- types_area_strata %>%
  mutate(n_infinite_flanders = n_type_flanders * sample_type,
         n_infinite_sac = n_type_sac * sample_type,
         n_infinite_subtype = n_subtype_flanders * extra_subtype,
         N = area_type / (32 * 32 /10000),
         N_sac = area_type_sac/ (32 * 32 /10000),
         N_subtype = area_subtype/(32 * 32 /10000),
         n_finite_flanders_tot = N * n_infinite_flanders/ (n_infinite_flanders + N - 1),
         n_finite_flanders = N * n_infinite_flanders/ (n_infinite_flanders + N - 1) * area_subtype_sac/area_type,
         n_finite_sac_tot = N_sac * n_infinite_sac/ (n_infinite_sac + N_sac - 1) * sac,
         n_finite_sac = N_sac * n_infinite_sac/ (n_infinite_sac + N_sac - 1) * area_subtype_sac/area_type_sac * sac,
         n_finite_subtype_tot =  N_subtype * n_infinite_subtype/ (n_infinite_subtype + N_subtype - 1),
         n_finite_subtype = N_subtype * n_infinite_subtype/ (n_infinite_subtype + N_subtype - 1) * area_subtype_sac/area_subtype,
         n_extra_sac = pmax(0, n_finite_sac - n_finite_flanders),
         n_extra_subtype = pmax(0, n_finite_subtype - n_finite_flanders - n_extra_sac))

main_type_samplesize_summary <- types_samplesize_detailed %>%
  group_by(main_type) %>%
  summarise(n_finite_flanders = ceiling(sum(n_finite_flanders)),
        n_extra_sac = ceiling(sum(n_extra_sac)),
        n_extra_subtypes = ceiling(sum(n_extra_subtype))) %>%
  ungroup() %>%
  mutate(n_tot = n_finite_flanders + n_extra_sac + n_extra_subtypes)
  
type_samplesize_summary <- types_samplesize_detailed %>%
  group_by(type) %>%
  summarise(n_finite_flanders = ceiling(sum(n_finite_flanders)),
        n_extra_sac = ceiling(sum(n_extra_sac)),
        n_extra_subtypes = ceiling(sum(n_extra_subtype))) %>%
  ungroup() %>%
  mutate(n_tot = n_finite_flanders + n_extra_sac + n_extra_subtypes)

total_samplesize <- types_samplesize_detailed %>%
  summarise(n_alltypes_flanders = ceiling(sum(n_finite_flanders)),
        n_alltypes_extra_sac = ceiling(sum(n_extra_sac)),
        n_alltypes_extra_subtypes = ceiling(sum(n_extra_subtype))) %>%
  mutate(n_alltypes_tot = n_alltypes_flanders + n_alltypes_extra_sac + n_alltypes_extra_subtypes)

result <- list(
  samplesize_detailed = types_samplesize_detailed,
  summary_types = type_samplesize_summary,
  summary_main_types = main_type_samplesize_summary,
  total = total_samplesize
)
return(result)
  
}
```


```{r}

types_area_strata_nonforest <- types_area_strata %>%
  filter(typeclass != "FW") %>%
  #filter(!main_type %in% c("1310", "5130")) %>%
  filter(str_sub(main_type, 1, 1) != "9")

types_area_strata_forest <- types_area_strata %>%
  filter(str_sub(main_type, 1, 1) == "9")

# original scenario

samplesize_original_nonforest <- calc_finitsamplesize_terrhab(types_area_strata = types_area_strata_nonforest)

samplesize_original_nonforest_subtype <- samplesize_original_nonforest$summary_types %>%
  mutate(type_cat = "nonforest type",
         scenario = "original",
         cycle_years = 12,
         n_flanders = 80,
         n_sac = 170,
         n_subtype = 80)
  
samplesize_original_forest <- calc_finitsamplesize_terrhab(types_area_strata = types_area_strata_forest, n_type_sac = 0)

samplesize_original_forest_subtype <- samplesize_original_forest$summary_types %>%
  mutate(type_cat = "forest type",
         scenario = "original",
         cycle_years = 12,
         n_flanders = 80,
         n_sac = 0,
         n_subtype = 80)

#scenario 1

samplesize_sc1_nonforest <- (calc_finitsamplesize_terrhab(types_area_strata = types_area_strata_nonforest, n_type_sac = 80))$summary_types %>%
  mutate(type_cat = "nonforest type",
         scenario = "alternative 1",
         cycle_years = 6,
         n_flanders = 80,
         n_sac = 80,
         n_subtype = 80)

#scenario 2

samplesize_sc2_nonforest <- (calc_finitsamplesize_terrhab(types_area_strata = types_area_strata_nonforest, n_type_sac = 80, n_type_flanders = 40, n_subtype_flanders = 40))$summary_types %>%
  mutate(type_cat = "nonforest type",
         scenario = "alternative 2",
         cycle_years = 6,
         n_flanders = 40,
         n_sac = 80,
         n_subtype = 40)

#scenario 3

samplesize_sc3_nonforest <- (calc_finitsamplesize_terrhab(types_area_strata = types_area_strata_nonforest, n_type_sac = 80, n_type_flanders = 80, n_subtype_flanders = 40))$summary_types %>%
  mutate(type_cat = "nonforest type",
         scenario = "alternative 3",
         cycle_years = 6,
         n_flanders = 80,
         n_sac = 80,
         n_subtype = 40)

#scenario 4

samplesize_sc4_nonforest <- (calc_finitsamplesize_terrhab(types_area_strata = types_area_strata_nonforest, n_type_sac = 80, n_type_flanders = 40, n_subtype_flanders = 80))$summary_types %>%
  mutate(type_cat = "nonforest type",
         scenario = "alternative 4",
         cycle_years = 6,
         n_flanders = 40,
         n_sac = 80,
         n_subtype = 80)



scenarios <- bind_rows(samplesize_original_forest_subtype,
                       samplesize_original_nonforest_subtype,
                       samplesize_sc1_nonforest,
                       samplesize_sc2_nonforest,
                       samplesize_sc3_nonforest,
                       samplesize_sc4_nonforest
                       ) %>%
  select(type_cat, scenario, n_flanders, n_sac, n_subtype, cycle_years, everything()) %>%
  left_join(types, by = "type") %>%
  filter(main_type != "6430") %>%
  mutate(fieldwork_team = ifelse(type_cat == "forest type", "anb",
                                 ifelse(main_type %in% c("2310", "2330", "4010", "4030", "6510"), "anb", "inbo")))


```

```{r}
samplesize_original_detailed <- bind_rows(samplesize_original_nonforest$samplesize_detailed,
                                          samplesize_original_forest$samplesize_detailed
) 

#write_vc(samplesize_original_detailed, root = "../output", file = "samplesize_mhq_terr_v2020", sorting = c("type", "sac"), strict = FALSE)

meetnet_selection <- samplesize_original_detailed %>%
  distinct(type, main_type, typelevel, n_polygons, area_type, prop_type_insac, prop_type, sample_type, area_subtype, n_polygons_subtype, prop_subtype, extra_subtype) %>%
  mutate(sample_type = ifelse(sample_type, "ja", "nee"),
         extra_subtype = ifelse(extra_subtype, "ja", "nee")) %>%
  group_by(main_type) %>%
  mutate(has_subtype = n_distinct(type) > 1) %>%
  ungroup() %>%
  mutate(subtype = ifelse(typelevel == "subtype", as.character(type), NA),
         extra_subtype = ifelse(typelevel == "subtype", extra_subtype, NA)) %>%
  filter(!(has_subtype & is.na(subtype))) %>%
  mutate(area_type = round(area_type, 1),
         area_subtype = round(area_subtype,1),
         prop_type_insac = round(prop_type_insac, 2),
         prop_type = round(prop_type, 2),
         prop_subtype = round(prop_subtype, 3)) %>%
  select(main_type, sample_type, n_polygons, area_type, prop_type_insac, prop_type, subtype, area_subtype, n_polygons_subtype, prop_subtype, extra_subtype)

```

Table \@ref(tab:tabscenario) compares the original design of @westra2014 with three alternative scenario's for open terrestrial habitats. For forest habitat types no alternative scenario's are considered since the monitoring scheme was designed to make it compatible with the Flemish Forest Inventory. The table shows the total sample size for the monitoring cycle and the total sample size per year after applying the finite population correction. The table also gives information on the total number of sampling units initially selected in Flanders ($tot_{fl}$) and the total number of extra sampling units selected in the SAC ($extra_{sac}$) and in the habitat subtypes ($extra_{subtype}$).       

```{r tabscenario}

scenarios_main_type <- scenarios %>%
  filter(n_finite_flanders > 0) %>%
  group_by(fieldwork_team, type_cat, main_type, scenario, n_flanders, n_sac, n_subtype, cycle_years) %>%
  summarise(n_finite_flanders = sum(n_finite_flanders),
            n_extra_sac = sum(n_extra_sac),
            n_extra_subtypes = sum(n_extra_subtypes),
            type_list = str_c(unique(main_type), collapse = ", ")) %>%
  ungroup() %>%
  mutate(n_total = n_finite_flanders + n_extra_sac + n_extra_subtypes,
         n_total_year = round(n_total/ cycle_years, 1),
         scenario = factor(scenario, levels = c("original", "alternative 1", "alternative 2", "alternative 3", "alternative 4"))) %>%
  arrange(fieldwork_team, scenario)

scenarios_total <- scenarios %>%
  filter(n_finite_flanders > 0) %>%
  group_by(type_cat, scenario, n_flanders, n_sac, n_subtype, cycle_years) %>%
  summarise(n_finite_flanders = sum(n_finite_flanders),
            n_extra_sac = sum(n_extra_sac),
            n_extra_subtypes = sum(n_extra_subtypes),
            type_list = str_c(unique(main_type), collapse = ", ")) %>%
  ungroup() %>%
  mutate(n_total = n_finite_flanders + n_extra_sac + n_extra_subtypes,
         n_total_year = ceiling(n_total/ cycle_years),
         scenario = factor(scenario, levels = c("original", "alternative 1", "alternative 2", "alternative 3", "alternative 4"))) %>%
  arrange(scenario)

scenarios_total %>%
  select("category" = type_cat, scenario, "cycle (years)" = cycle_years, "$n_{fl}$"  = n_flanders, "$n_{sac}$" = n_sac, "$n_{subtype}$" = n_subtype, "$tot_{fl}$" =  n_finite_flanders, "$extra_{sac}$" =  n_extra_sac, "$extra_{subtype}$" =  n_extra_subtypes, "total per cycle" = n_total, "total per year" = n_total_year) %>%
  arrange(desc(category)) %>%
  kable(booktabs = TRUE,
        align = "c",
        escape = FALSE,
        caption = "Comparison between original design and three alternative scenario's. The selected scenario's are shown in bold.") %>%
  kable_styling() %>%
  collapse_rows(c(1), target = 1) %>%
  row_spec(c(4,6), bold = TRUE)

#scenarios_team %>%
#  write.csv2("../output/scenario_team.csv")
```


```{r}
samplesize_scenario3_nonforest <- calc_finitsamplesize_terrhab(types_area_strata = types_area_strata_nonforest, n_type_sac = 80, n_type_flanders = 80, n_subtype_flanders = 40)

samplesize_scenario3_detailed <- bind_rows(samplesize_scenario3_nonforest$samplesize_detailed,
                                          samplesize_original_forest$samplesize_detailed
) 

#write_vc(samplesize_scenario3_detailed, root = "../output", file = "samplesize_mhq_terr_v2021", sorting = c("type", "sac"), strict = FALSE)


```

By comparing the costs (the number of sampling units that have to be measured per year) and the benefits (the minimal differences for status and trend that can be detected), we decide to implement alternative 3 for the open habitat types. It allows the estimation of short-term trends (12 year) in habitat quality based on paired observations, as the monitoring cycle is lowered to 6 years. It also allows the estimation of the status of habitat quality on the scale of Flanders every six years with (approximately) the same precision as the original design, as the sample size per habitat type is kept to 80. The cost of alternative 3 is only slightly higher (11 extra sampling units per year) compared to the original design.    

Annex \@ref(annex3) gives an overview of the sample size per habitat type and subtype.

