# Sample design for standing water habitat types 

## Original design

The original design for the standing water habitat types in @westra2014 consists of a stratified sample with a combination of following strata:

+ area class of the water surface
  + `Opp <= 1 ha`
  + `1 ha < Opp <= 5 ha`
  + `5 ha < Opp < 50 ha`
  + `>= 50 ha`
+ inside or outside a special area of conservation (SAC)

The original design can be summarized as follows:

+ draw a sample for 3130_aom, 3150, 3160 in area classes `Opp <= 1 ha` and  `1 ha < Opp <= 5 ha`
  + with an initial sample size of 40 sampling units for Flanders ($n_{fl}$)
  + select extra sampling units within SAC until 80 sampling units are selected
  + apply the finite population correction (see paragraph \@ref(scenarioterr))
  + divide sample size equally over both area classes
+ selection of all water surfaces for other area classes and other habitat types and habitat subtypes

Although a monitoring cycle of 12 years was proposed in @westra2014, the monitoring schemes for standing water habitat types were implemented with a monitoring cycle of 6 years. This is because the field work could be combined with other surveys. To be able to detect short-term trends (12 years) in habitat quality we will keep this monitoring cycle of 6 years. 

## Impact of the sample frame on the sample size in the original design

The number of water surfaces within each stratum will have an impact on the number of selected sampling units.
The main impact is that all water surfaces in the area classes `5 ha < Opp < 50 ha` and `>= 50 ha` are selected in the original design. Furthermore, the number of polygons also has an impact on the finite population correction.

Therefore, we want to compare the sampling frame that was originally used for designing the monitoring schemes with the latest available sampling frame.  

The original sample frame consists of the polygons with standing water habitat types in the 2014 version of the Flemish habitat map. 

```{r}

watersurfaces_hab_polygons_2014_orig <- read_sf("data/watersurfaces_2014/BWK_plassen_versie 20140324.shp")

watersurfaces_hab_polygons_2014 <- watersurfaces_hab_polygons_2014_orig %>%
  mutate(sac = (SBZH == 1) & (!is.na(SBZH))) %>%
  select(pol_id = Pol_ID, sac)

check <- nrow(watersurfaces_hab_polygons_2014) == n_distinct(watersurfaces_hab_polygons_2014$pol_id)

watersurfaces_hab_types_2014_orig <- read.table("data/watersurfaces_2014/BWK_hab_aq.txt", header=TRUE)

watersurfaces_hab_types_2014 <- watersurfaces_hab_types_2014_orig %>%
  select(pol_id = Pol_ID, shape_area = Shape_Area, code, phab) %>%
  filter(str_detect(code, "gh", negate = TRUE)) %>%
  mutate(code = ifelse(code == "3130", "3130_aom", code),
         patch_area = shape_area * phab/100,
         area_ha = patch_area/10000,
         area_polygon_ha = shape_area/10000,
         area_class = ifelse(area_ha <= 1, "Opp <= 1 ha",
                             ifelse(area_ha <= 5, "1 ha < Opp <= 5 ha",
                                    ifelse(area_ha < 50, "5 ha < Opp < 50 ha", ">= 50 ha"))),
         area_class = factor(area_class, levels = c("Opp <= 1 ha", "1 ha < Opp <= 5 ha", "5 ha < Opp < 50 ha", ">= 50 ha"))) %>%
  left_join(st_drop_geometry(watersurfaces_hab_polygons_2014), by = "pol_id") %>%
  rename(type = code) %>%
  mutate(version = "habitatmap_2014")

```


```{r}

sac_sf <- read_admin_areas(dsn = "sac") %>%
  group_by(sac_name) %>%
  summarise(n_pol = n()) %>%
  ungroup() %>%
  select(sac_name)

```

Since 2014 new data sources became available:

+ the map of the water surfaces in Flanders [@inbo2020_2]
+ updated information on the presence of aquatic habitat types, made available through the 2020 version of the Flemish habitat map [@inbo2020]

Both data sources are combined in the map of standing water habitat types [@westra2021_2], which is used as the current sample frame.   

```{r}

watersurfaces_populationunits <- read_vc(root = "data", file = "mhq_watersurfaces_populationunits")

watersurfaces_hab <- read_watersurfaces_hab()

filepath <-  file.path(fileman_up("n2khab_data"), "20_processed/watersurfaces_hab/watersurfaces_hab.gpkg")

check_version <- n2khab::xxh64sum(filepath) == "5792b496a94d0524"

watersurfaces_hab <- watersurfaces_hab$watersurfaces_polygons %>%
  semi_join(watersurfaces_populationunits, by = "polygon_id") %>%
  select(polygon_id)

watersurfaces_extra <- read_watersurfaces() %>%
  filter(polygon_id %in% watersurfaces_populationunits$polygon_id) %>%
  filter(! polygon_id %in% watersurfaces_hab$polygon_id) %>%
  select(polygon_id)

watersurfaces_area_sf <- watersurfaces_hab %>%
  bind_rows(watersurfaces_extra) %>%
  mutate(area_ha = drop_units(st_area(geom))/10000,
         area_polygon_ha = area_ha,
         area_class = ifelse(area_ha <= 1, "Opp <= 1 ha",
                             ifelse(area_ha <= 5, "1 ha < Opp <= 5 ha",
                                    ifelse(area_ha < 50, "5 ha < Opp < 50 ha", ">= 50 ha"))),
         area_class = factor(area_class, levels = c("Opp <= 1 ha", "1 ha < Opp <= 5 ha", "5 ha < Opp < 50 ha", ">= 50 ha"))) %>%
  st_join(sac_sf) %>%
  mutate(sac = !is.na(sac_name)) 

watersurfaces_area <- watersurfaces_area_sf %>%
  st_drop_geometry()

watersurfaces_types_2021 <- watersurfaces_populationunits %>%
  left_join(watersurfaces_area, by = "polygon_id") %>%
  mutate(type = ifelse(type == "3130", "3130_aom", as.character(type))) %>%
  mutate(version = "watersurfaces_hab_2021")

watersurfaces_types <- watersurfaces_types_2021 %>%
  bind_rows(watersurfaces_hab_types_2014)

```

```{r}
watersurfaces_hab_distribution <- watersurfaces_types %>%
  group_by(version, type, sac, area_class) %>%
  summarise(n_watersurfaces = n(),
            area_ha = sum(area_ha)) %>%
  ungroup() %>%
  group_by(version, type, sac) %>%
  mutate(total_n_sac = sum(n_watersurfaces),
         total_area_sac = sum(area_ha)) %>%
  ungroup() 

```



```{r}
n_fland <- 40
n_sac <- 80

samplesize <- watersurfaces_hab_distribution %>%
  group_by(version, type, area_class) %>%
  mutate(total_area_class = sum(n_watersurfaces)) %>%
  ungroup() %>%
  group_by(version, type) %>%
  mutate(total = sum(n_watersurfaces)) %>%
  ungroup() %>%
  mutate(take_sample = total >= 50 & area_class %in% c("Opp <= 1 ha", "1 ha < Opp <= 5 ha")) %>%
  group_by(version, type, take_sample) %>%
  mutate(n_watersurfaces_sample = sum(n_watersurfaces * take_sample)) %>%
  ungroup() %>% 
  group_by(version, type, take_sample, sac) %>%
  mutate(n_watersurfaces_sample_sac = sum(n_watersurfaces * take_sample)) %>%
  ungroup() %>%
  mutate(n_finite_fland = n_watersurfaces_sample * n_fland / (n_fland + (n_watersurfaces_sample - 1)),
         n_finite_sac = n_watersurfaces_sample_sac * n_sac / (n_sac + (n_watersurfaces_sample_sac - 1)),
         prop_sac = n_watersurfaces_sample_sac/n_watersurfaces_sample, 
         samplesize = ifelse(!take_sample, n_watersurfaces,
                             ifelse(sac, n_finite_sac/2, n_finite_fland * prop_sac/2)),
         samplesize = pmin(samplesize, n_watersurfaces))

```

Table \@ref(tab:tabtotalsz) shows the total sample size and the total number of sampling units in the sample frame, for both the original sample frame (habitatmap_2014) and the current sample frame (watersurfaces_hab_2021). It is clear that the watersurfaces_hab_2021 sample frame contains a lot more sampling units than habitatmap_2014, which results in a much higher sample size. In Table \@ref(tab:tabsamplesizerepro) the comparison is shown per habitat type and subtype.

```{r samplesizetotal}
samplesize_total <- samplesize %>%
  filter(version %in% c("habitatmap_2014", "watersurfaces_hab_2021")) %>%
  group_by(version, area_class) %>%
  summarise(samplesize = sum(samplesize),
            n_watersurfaces = sum(n_watersurfaces)) %>%
  ungroup() %>%
  mutate(samplesize_show = str_c(round(samplesize), "/", n_watersurfaces),
         .keep = "unused") %>%
  spread(key = area_class, value = samplesize_show, fill = 0) %>%
  arrange(version) 
```

```{r tabtotalsz}
samplesize_total %>%
  kable(caption = "Total sample size per area class for standing water habitat types based on the original design / the total number of sampling units for both the original sample frame (habitatmap_2014) and the current sample frame (watersurfaces_hab_2021)") %>%
  kable_styling()
```


```{r tabsamplesizerepro}

samplesize_overview <- samplesize %>%
  filter(version %in% c("habitatmap_2014", "watersurfaces_hab_2021")) %>%
  group_by(type, version, area_class) %>%
  summarise(samplesize = sum(samplesize), 
            n_watersurfaces = sum(n_watersurfaces)) %>%
  ungroup() %>%
  mutate(samplesize_show = str_c(round(samplesize), "/", n_watersurfaces),
         .keep = "unused") %>%
  spread(key = area_class, value = samplesize_show, fill = 0) %>%
  arrange(type, version) 

samplesize_overview %>%
  kable(caption = "Total sample size per area class for standing water habitat types based on the original design / the total number of sampling units for both the original sample frame (habitatmap_2014) and the current sample frame (watersurfaces_hab_2021)") %>%
  kable_styling() %>%
  collapse_rows(c(1,2)) 
  
```






The total sample size almost doubles.
Given the available resources, the sample size that results from the current sample frame is not feasible.
Therefore we will explore some alternative scenarios.

## Alternative scenarios

We propose the following alternative scenario:

+ draw a sample for all types with more than 30 watersurfaces in the area classes `Opp <= 1 ha`, `1 ha < Opp <= 5 ha` and `5 ha < Opp <= 50 ha` 
  + sample size = 80 for Flanders
  + extra sampling units within sac until 80 sampling units are selected
  + minimum 40 sampling units for each subtype (select extra sampling units if needed)
  + finite population correction
  + distribute sample size over different area classes
    + using the log of the total area within each area class as weight (alternative 1)
    + using the square root of the total area within each area class as weight (alternative 2)
    + dividing the sample size equally over the three area classes (alternative 3)
+ selection of all watersurfaces for other area classes (`Opp > 50 ha`) and other habitat types (type 3110)



```{r}

samplesize_watersurfaces <- function(watersurfaces_hab_distribution,
                                     n_fl,
                                     n_sac,
                                     n_subtype) {
  
  samplesize_fl <- watersurfaces_hab_distribution %>%
  mutate(main_type = str_sub(type, 1, 4)) %>%
  group_by(version, main_type, area_class, sac) %>%
  summarise(area_ha_sac = sum(area_ha),
            n_watersurfaces_sac = sum(n_watersurfaces)) %>%
  ungroup() %>%
  group_by(version, main_type, area_class) %>%
  mutate(n_watersurfaces = sum(n_watersurfaces_sac),
         area_ha = sum(area_ha_sac)) %>%
  ungroup() %>%
  mutate(weight1 = log(area_ha + 1),
          weight2 = area_ha ^ (1/2)) %>%
  group_by(version, main_type, sac) %>%
  mutate(total_area_ha = sum(area_ha * (area_class != ">= 50 ha")),
         total_watersurfaces = sum(n_watersurfaces * (area_class != ">= 50 ha")),
         total_weight1 = sum(weight1 * (area_class != ">= 50 ha")),
         total_weight2 = sum(weight2 * (area_class != ">= 50 ha"))) %>%
  ungroup() %>%
  mutate(n_finite_fland = total_watersurfaces * n_fl / (n_fl + (total_watersurfaces - 1))) %>%
  mutate(samplesize_fl_alt1 = ifelse(area_class != ">= 50 ha", n_finite_fland * weight1 / total_weight1, n_watersurfaces),
         samplesize_fl_alt2 = ifelse(area_class != ">= 50 ha", n_finite_fland * weight2 / total_weight2, n_watersurfaces),
         samplesize_fl_alt3 = ifelse(area_class != ">= 50 ha", n_finite_fland /3, n_watersurfaces)) %>%
  mutate(samplesize_fl_alt1 = pmin(samplesize_fl_alt1 * n_watersurfaces_sac/n_watersurfaces, n_watersurfaces),
         samplesize_fl_alt2 = pmin(samplesize_fl_alt2 * n_watersurfaces_sac/n_watersurfaces, n_watersurfaces),
         samplesize_fl_alt3 = pmin(samplesize_fl_alt3 * n_watersurfaces_sac/n_watersurfaces, n_watersurfaces)) 


samplesize_sac <- watersurfaces_hab_distribution %>%
  filter(sac) %>%
  mutate(main_type = str_sub(type, 1, 4)) %>%
  group_by(version, sac, main_type, area_class) %>%
  summarise(area_ha = sum(area_ha),
            n_watersurfaces = sum(n_watersurfaces)) %>%
  ungroup() %>%
  mutate(weight1 = log(area_ha + 1),
          weight2 = area_ha ^ (1/2)) %>%
  group_by(version, main_type) %>%
  mutate(total_area_ha = sum(area_ha * (area_class != ">= 50 ha")),
         total_watersurfaces = sum(n_watersurfaces),
         total_weight1 = sum(weight1 * (area_class != ">= 50 ha")),
         total_weight2 = sum(weight2 * (area_class != ">= 50 ha"))) %>%
  ungroup() %>%
  mutate(n_finite_sac = total_watersurfaces * n_sac / (n_sac + (total_watersurfaces - 1))) %>%
  mutate(samplesize_sac_alt1 = ifelse(area_class != ">= 50 ha", n_finite_sac * weight1 / total_weight1, n_watersurfaces),
         samplesize_sac_alt2 = ifelse(area_class != ">= 50 ha", n_finite_sac * weight2 / total_weight2, n_watersurfaces),
         samplesize_sac_alt3 = ifelse(area_class != ">= 50 ha", n_finite_sac /3, n_watersurfaces)) %>%
  mutate(samplesize_sac_alt1 = pmin(samplesize_sac_alt1, n_watersurfaces),
         samplesize_sac_alt2 = pmin(samplesize_sac_alt2, n_watersurfaces),
         samplesize_sac_alt3 = pmin(samplesize_sac_alt3, n_watersurfaces))

samplesize_subtype <- watersurfaces_hab_distribution %>%
  mutate(main_type = str_sub(type, 1, 4)) %>%
  filter(main_type == "3130") %>%
  group_by(version, type, area_class, sac) %>%
  summarise(n_watersurfaces_sac = sum(n_watersurfaces),
            area_ha_sac = sum(area_ha)) %>%
  ungroup() %>%
  group_by(version, type, area_class) %>%
  mutate(n_watersurfaces = sum(n_watersurfaces_sac),
         area_ha = sum(area_ha_sac)) %>%
  ungroup() %>%
  mutate(weight1 = log(area_ha + 1),
          weight2 = area_ha ^ (1/2)) %>%
  group_by(version, type, sac) %>%
  mutate(total_area_ha = sum(area_ha * (area_class != ">= 50 ha")),
         total_watersurfaces = sum(n_watersurfaces),
         total_weight1 = sum(weight1 * (area_class != ">= 50 ha")),
         total_weight2 = sum(weight2 * (area_class != ">= 50 ha"))) %>%
  ungroup() %>%
  mutate(n_finite_subtype = total_watersurfaces * n_subtype / (n_subtype + (total_watersurfaces - 1))) %>%
  mutate(samplesize_subtype_alt1 = ifelse(area_class != ">= 50 ha", n_finite_subtype * weight1 / total_weight1, n_watersurfaces),
         samplesize_subtype_alt2 = ifelse(area_class != ">= 50 ha", n_finite_subtype * weight2 / total_weight2, n_watersurfaces),
         samplesize_subtype_alt3 = ifelse(area_class != ">= 50 ha", n_finite_subtype /3, n_watersurfaces)) %>%
  mutate(samplesize_subtype_alt1 = pmin(samplesize_subtype_alt1 * n_watersurfaces_sac/n_watersurfaces, n_watersurfaces_sac),
         samplesize_subtype_alt2 = pmin(samplesize_subtype_alt2 * n_watersurfaces_sac/n_watersurfaces, n_watersurfaces_sac),
         samplesize_subtype_alt3 = pmin(samplesize_subtype_alt3 * n_watersurfaces_sac/n_watersurfaces, n_watersurfaces_sac))

samplesize_alt <- watersurfaces_hab_distribution %>%
  mutate(main_type = str_sub(type, 1, 4),
         n_fl = n_fl,
         n_sac = n_sac,
         n_subtype = n_subtype) %>%
  select(version, n_fl, n_sac, n_subtype, main_type, type, sac, area_class, area_ha, n_watersurfaces) %>%
  group_by(version, main_type) %>%
  mutate(n_watersurfaces_sample = sum(n_watersurfaces)) %>%
  ungroup() %>%
  left_join(select(samplesize_fl, sac, version, main_type, area_class, samplesize_fl_alt1, samplesize_fl_alt2, samplesize_fl_alt3), by = c("version", "main_type", "area_class", "sac")) %>%
  left_join(select(samplesize_sac, sac, version, main_type, area_class, samplesize_sac_alt1, samplesize_sac_alt2, samplesize_sac_alt3), by = c("version", "main_type", "area_class", "sac")) %>%
  group_by(version, main_type, sac, area_class) %>%
  mutate(n_watersurfaces_main_type = sum(n_watersurfaces)) %>%
  ungroup() %>%
  mutate(samplesize_main_type_alt1 = ifelse(area_class == ">= 50 ha" | n_watersurfaces_sample < 30, n_watersurfaces,
                                            ifelse(sac, pmax(samplesize_sac_alt1, samplesize_fl_alt1), samplesize_fl_alt1)),
         samplesize_main_type_alt2 = ifelse(area_class == ">= 50 ha" | n_watersurfaces_sample < 30, n_watersurfaces,
                                            ifelse(sac, pmax(samplesize_sac_alt2, samplesize_fl_alt2), samplesize_fl_alt2)),
         samplesize_main_type_alt3 = ifelse(area_class == ">= 50 ha" | n_watersurfaces_sample < 30, n_watersurfaces,
                                            ifelse(sac, pmax(samplesize_sac_alt3, samplesize_fl_alt3), samplesize_fl_alt3)),
         samplesize_type_alt1 = samplesize_main_type_alt1 * n_watersurfaces / n_watersurfaces_main_type,
         samplesize_type_alt2 = samplesize_main_type_alt2 * n_watersurfaces / n_watersurfaces_main_type,
         samplesize_type_alt3 = samplesize_main_type_alt3 * n_watersurfaces / n_watersurfaces_main_type,
         ) %>%
  left_join(select(samplesize_subtype, sac, version, type, area_class, samplesize_subtype_alt1, samplesize_subtype_alt2, samplesize_subtype_alt3), by = c("version", "type", "area_class", "sac")) %>%
  mutate(samplesize_total_alt1 = round(pmax(samplesize_type_alt1, samplesize_subtype_alt1, na.rm = TRUE)),
         samplesize_total_alt2 = round(pmax(samplesize_type_alt2, samplesize_subtype_alt2, na.rm = TRUE)),
         samplesize_total_alt3 = round(pmax(samplesize_type_alt3, samplesize_subtype_alt3, na.rm = TRUE)))

  return(samplesize_alt)

}
  

samplesize_alt1 <- samplesize_watersurfaces(watersurfaces_hab_distribution = watersurfaces_hab_distribution,
                                            n_fl = 80, n_sac = 80, n_subtype = 40) %>%
  mutate(scenario = "alternative 1")

samplesize_alt2 <- samplesize_watersurfaces(watersurfaces_hab_distribution = watersurfaces_hab_distribution,
                                            n_fl = 80, n_sac = 0, n_subtype = 40) %>%
  mutate(scenario = "alternative 2")


```

```{r}
samplesize_alt_total <- samplesize_alt1 %>%
  bind_rows(samplesize_alt2) %>%
  filter(type != "2190_a") %>%
  select(version, scenario, n_fl, n_sac, n_subtype, type, area_class, sac, n_watersurfaces, samplesize = samplesize_total_alt3) %>% 
  group_by(scenario, n_fl, n_sac, n_subtype, version, area_class) %>%
  summarise(samplesize = sum(samplesize),
            n_watersurfaces = sum(n_watersurfaces)) %>%
  ungroup() %>%
  mutate(samplesize_show = str_c(round(samplesize), "/", n_watersurfaces),
         .keep = "unused") %>%
  spread(key = area_class, value = samplesize_show, fill = 0)
```

Table \@ref(tab:alttotalws) shows the total sample sizes per area class for the different alternative scenarios. Table \@ref(tab:alttypews) shows the sample sizes per type, per area class and inside/outside SAC for the different alternative scenarios.

```{r alttotalws}
samplesize_alt_total %>%
  filter(version == "watersurfaces_hab_2021") %>%
  select(-version) %>%
  kable(caption = "Alternative sample size scenarios compared the to original design: total number of sampling units/ total number of watersurfaces") %>%
  kable_styling()
```


```{r alttypews}
samplesize_alt_overview <- samplesize_alt1 %>%
  bind_rows(samplesize_alt2) %>%
  select(version, scenario, n_fl, n_sac, n_subtype, type, area_class, sac, n_watersurfaces, samplesize = samplesize_total_alt3) %>%
  filter(version %in% c("watersurfaces_hab_2021")) %>%
  # filter(type != "2190_a") %>%
  mutate(samplesize_show = str_c(round(samplesize), "/", n_watersurfaces),
         .keep = "unused") %>%
  spread(key = area_class, value = samplesize_show, fill = 0) %>%
  select(type, version, scenario, everything()) 

samplesize_alt_overview %>%
  arrange(type, scenario) %>%
  select(-version, -n_fl, -n_sac, -n_subtype) %>%
  mutate(sac = ifelse(sac, "inside", "outside")) %>%
  kable(caption = "Alternative sample size scenarios compared the to original design: number of sampling units per type/ total number of watersurfaces per type") %>%
  kable_styling() %>%
  collapse_rows(c(1,2, 3), target = 1)
  
```

Sample sizes of the alternative scenarios are slightly higher compared to the original sample size (except for 3130_aom).
However, this increase is smaller compared to the case where the original sample size calculation is applied to watersurfaces_hab.

