# Sample design for standing water habitat types {#scenariostandwater}

## Standing water habitat types included in MHQ

A monitoring scheme for all standing water habit types and subtypes are included in MHQ. These are all the habitat types for which the code in Annex \@ref(annex1) starts with 31, but also the habitat subtype 2190_a (Dune slack ponds). Habitat subtype 2190_a was overlooked in @westra2014, however a monitoring scheme was designed and implemented in 2016.

## Original design

The population unit for the standing water monitoring scheme is an entire water body. The original design for the standing water habitat types in @westra2014 consists of a stratified sample with a combination of following strata:

*   area class of the water body
    +   `area <= 1 ha`
    +   `1 ha < area <= 5 ha`
    +   `5 ha < area < 50 ha`
    +   `area >= 50 ha`
*   inside or outside a special area of conservation (SAC)

The original design can be summarized as follows:

*   draw a sample for 3130_aom, 3150, 3160 in area classes `area <= 1 ha` and `1 ha < area <= 5 ha`
    +   with an initial sample size of 40 sampling units for Flanders ($n_{fl}$)
    +   select extra sampling units within SAC until 80 sampling units are selected
    +   apply the finite population correction (see paragraph \@ref(scenarioterr))
    +   divide sample size equally over both area classes
*   selection of all water bodies for other area classes and other habitat types and habitat subtypes

Although a monitoring cycle of 12 years was proposed in @westra2014, the monitoring schemes for standing water habitat types were implemented with a monitoring cycle of 6 years. This is because the field work could be combined with other surveys. To be able to detect short-term trends (12 years) in habitat quality we will keep this monitoring cycle of 6 years.

## Impact of the sample frame on the sample size in the original design

The number of water bodies within each stratum will have an impact on the number of selected sampling units. The main impact is that all water bodies in the area classes `5 ha < area < 50 ha` and `area >= 50 ha` are selected in the original design. Furthermore, the number of population units also has an impact on the finite population correction.

Therefore, we want to compare the sampling frame that was originally used for designing the monitoring schemes with the latest available sampling frame.

The original sample frame consists of the polygons with standing water habitat types in the 2014 version of the Natura 2000 habitat map.

```{r}

watersurfaces_hab_polygons_2014_orig <- read_sf("data/watersurfaces_2014/BWK_plassen_versie 20140324.shp")

watersurfaces_hab_polygons_2014 <- watersurfaces_hab_polygons_2014_orig %>%
  mutate(sac = (SBZH == 1) & (!is.na(SBZH))) %>%
  select(pol_id = Pol_ID, sac)

check <- nrow(watersurfaces_hab_polygons_2014) == n_distinct(watersurfaces_hab_polygons_2014$pol_id)

watersurfaces_hab_types_2014_orig <- read.table("data/watersurfaces_2014/BWK_hab_aq.txt", header=TRUE)

watersurfaces_hab_types_2014 <- watersurfaces_hab_types_2014_orig %>%
  select(pol_id = Pol_ID, shape_area = Shape_Area, code, phab) %>%
  filter(str_detect(code, "gh", negate = TRUE)) %>%
  mutate(code = ifelse(code == "3130", "3130_aom", code),
         patch_area = shape_area * phab/100,
         area_ha = patch_area/10000,
         area_polygon_ha = shape_area/10000,
         area_class = ifelse(area_ha <= 1, "area <= 1 ha",
                             ifelse(area_ha <= 5, "1 ha < area <= 5 ha",
                                    ifelse(area_ha < 50, "5 ha < area < 50 ha", "area >= 50 ha"))),
         area_class = factor(area_class, levels = c("area <= 1 ha", "1 ha < area <= 5 ha", "5 ha < area < 50 ha", "area >= 50 ha"))) %>%
  left_join(st_drop_geometry(watersurfaces_hab_polygons_2014), by = "pol_id") %>%
  rename(type = code) %>%
  mutate(version = "habitatmap_2014")

```

```{r}

sac_sf <- read_admin_areas(dsn = "sac") %>%
  group_by(sac_name) %>%
  summarise(n_pol = n()) %>%
  ungroup() %>%
  select(sac_name)

```

Since 2014 new data sources became available:

-   the map of the water surfaces in Flanders [@inbo2020_2]
-   updated information on the presence of aquatic habitat types, made available through the 2020 version of the Natura 2000 habitat map [@inbo2020]

Both data sources are combined in the map of standing water habitat types [@westra2021_2], which is used as the current sample frame. It can be imported in R by using the function `read_watersurfaces_hab()` of the n2khab package [@vanderhaeghe2022].

```{r}

watersurfaces_populationunits <- read_vc(root = "data", file = "mhq_watersurfaces_populationunits")

watersurfaces_hab <- read_watersurfaces_hab()

filepath <-  file.path(fileman_up("n2khab_data"), "20_processed/watersurfaces_hab/watersurfaces_hab.gpkg")

check_version <- n2khab::xxh64sum(filepath) == "5792b496a94d0524"

watersurfaces_hab <- watersurfaces_hab$watersurfaces_polygons %>%
  semi_join(watersurfaces_populationunits, by = "polygon_id") %>%
  select(polygon_id)

watersurfaces_extra <- read_watersurfaces() %>%
  filter(polygon_id %in% watersurfaces_populationunits$polygon_id) %>%
  filter(! polygon_id %in% watersurfaces_hab$polygon_id) %>%
  select(polygon_id)

watersurfaces_area_sf <- watersurfaces_hab %>%
  bind_rows(watersurfaces_extra) %>%
  mutate(area_ha = drop_units(st_area(geom))/10000,
         area_polygon_ha = area_ha,
         area_class = ifelse(area_ha <= 1, "area <= 1 ha",
                             ifelse(area_ha <= 5, "1 ha < area <= 5 ha",
                                    ifelse(area_ha < 50, "5 ha < area < 50 ha", "area >= 50 ha"))),
         area_class = factor(area_class, levels = c("area <= 1 ha", "1 ha < area <= 5 ha", "5 ha < area < 50 ha", "area >= 50 ha"))) %>%
  st_join(sac_sf) %>%
  mutate(sac = !is.na(sac_name)) 

watersurfaces_area <- watersurfaces_area_sf %>%
  st_drop_geometry()

watersurfaces_types_2021 <- watersurfaces_populationunits %>%
  left_join(watersurfaces_area, by = "polygon_id") %>%
  mutate(type = ifelse(type == "3130", "3130_aom", as.character(type))) %>%
  mutate(version = "watersurfaces_hab_2021")

watersurfaces_types <- watersurfaces_types_2021 %>%
  bind_rows(watersurfaces_hab_types_2014)

```

```{r}
watersurfaces_hab_distribution <- watersurfaces_types %>%
  group_by(version, type, sac, area_class) %>%
  summarise(n_watersurfaces = n(),
            area_ha = sum(area_ha)) %>%
  ungroup() %>%
  group_by(version, type, sac) %>%
  mutate(total_n_sac = sum(n_watersurfaces),
         total_area_sac = sum(area_ha)) %>%
  ungroup() 

```

```{r}
n_fland <- 40
n_sac <- 80

samplesize <- watersurfaces_hab_distribution %>%
  group_by(version, type, area_class) %>%
  mutate(total_area_class = sum(n_watersurfaces)) %>%
  ungroup() %>%
  group_by(version, type) %>%
  mutate(total = sum(n_watersurfaces)) %>%
  ungroup() %>%
  mutate(take_sample = total >= 50 & area_class %in% c("area <= 1 ha", "1 ha < area <= 5 ha")) %>%
  group_by(version, type, take_sample) %>%
  mutate(n_watersurfaces_sample = sum(n_watersurfaces * take_sample)) %>%
  ungroup() %>% 
  group_by(version, type, take_sample, sac) %>%
  mutate(n_watersurfaces_sample_sac = sum(n_watersurfaces * take_sample)) %>%
  ungroup() %>%
  mutate(n_finite_fland = n_watersurfaces_sample * n_fland / (n_fland + (n_watersurfaces_sample - 1)),
         n_finite_sac = n_watersurfaces_sample_sac * n_sac / (n_sac + (n_watersurfaces_sample_sac - 1)),
         prop_sac = n_watersurfaces_sample_sac/n_watersurfaces_sample, 
         samplesize = ifelse(!take_sample, n_watersurfaces,
                             ifelse(sac, n_finite_sac/2, n_finite_fland * prop_sac/2)),
         samplesize = pmin(samplesize, n_watersurfaces))

```

Table \@ref(tab:tabtotalsz) shows the total sample size and the total number of sampling units in the sample frame, for both the original sample frame (habitatmap_2014) and the current sample frame (watersurfaces_hab_2021). It is clear that the watersurfaces_hab_2021 sample frame contains a lot more sampling units than the original sample frame (habitatmap_2014), which results in a much higher sample size. In Table \@ref(tab:tabsamplesizerepro) the comparison is shown per habitat type and subtype. The current sample frame also contains the 2190_a habitat subtype resulting in the selection of sampling units for this subtype. This is not the case for the original sample frame.

```{r samplesizetotal}
samplesize_total <- samplesize %>%
  filter(version %in% c("habitatmap_2014", "watersurfaces_hab_2021")) %>%
  group_by(version, area_class) %>%
  summarise(samplesize = sum(samplesize),
            n_watersurfaces = sum(n_watersurfaces)) %>%
  ungroup() %>%
  mutate(samplesize_show = str_c(round(samplesize), " of ", n_watersurfaces),
         .keep = "unused") %>%
  spread(key = area_class, value = samplesize_show, fill = 0) %>%
  arrange(version) %>%
  mutate(scenario = "original")
```

```{r tabtotalsz, eval = TRUE}
samplesize_total %>%
  mutate(version = ifelse(version == "habitatmap_2014", "habitatmap\\_2014", "watersurfaces\\_hab\\_2021")) %>%
  rename("version sample frame" = version) %>%
  select(-scenario) %>%
  kable(caption = "Total sample size per area class for standing water habitat types based on the original design and the total number of sampling units for both the original sample frame (habitatmap\\_2014) and the current sample frame (watersurfaces\\_hab\\_2021)",
        escape = FALSE,
        booktabs = TRUE, 
        format = format_table) %>%
  kable_styling()
```

```{r tabsamplesizerepro, eval = TRUE}

samplesize_overview <- samplesize %>%
  filter(version %in% c("habitatmap_2014", "watersurfaces_hab_2021")) %>%
  group_by(type, version, area_class) %>%
  summarise(samplesize = sum(samplesize), 
            n_watersurfaces = sum(n_watersurfaces)) %>%
  ungroup() %>%
  mutate(samplesize_show = str_c(round(samplesize), " of ", n_watersurfaces),
         .keep = "unused") %>%
  spread(key = area_class, value = samplesize_show, fill = 0) %>%
  arrange(type, version) 

samplesize_overview %>%
  mutate(version = ifelse(version == "habitatmap_2014", "habitatmap\\_2014", "watersurfaces\\_hab\\_2021"),
         type = ifelse(type == "2190_a", "2190\\_a",
                       ifelse(type == "3130_aom", "3130\\_aom", 
                              ifelse(type == "3130_na", "3130\\_na", type)))) %>%
  rename("version sample frame" = version) %>%
  kable(caption = "Total sample size and the total number of sampling units per standing water habitat type and per area class based on the original design for both the original sample frame (habitatmap\\_2014) and the current sample frame (watersurfaces\\_hab\\_2021)",
        booktabs = TRUE,
        escape = FALSE, 
        format = format_table) %>%
  kable_styling(latex_options = "scale_down") %>%
  column_spec(c(3,4,5,6), width = "2cm") %>%
  collapse_rows(c(1,2))
  
```

The total sample size almost doubles. Since the available resources for field work has not increased, the sample size that results from the current sample frame is not feasible. Therefore we will explore two alternative scenarios.

## Alternative scenarios

We propose the following alternative scenarios:

+ draw a sample for all types with more than 30 water surfaces in the area classes `area <= 1 ha`, `1 ha < area <= 5 ha` and `5 ha < area <= 50 ha`
  + select an initial sample size of 80 for Flanders
  + select extra sampling units within SAC
    + alternative 1: until 80 sampling units are selected
    + alternative 2: no extra sampling units
  + select minimum 40 sampling units for each subtype (select extra sampling units if needed)
  + finite population correction
  + distribute sample size over different area classes dividing the sample size equally over the three area classes
+ selection of all water surfaces for other area classes (`area > 50 ha`) and other habitat types (type 3110)

```{r}

samplesize_watersurfaces <- function(watersurfaces_hab_distribution,
                                     n_fl,
                                     n_sac,
                                     n_subtype) {
  
  samplesize_fl <- watersurfaces_hab_distribution %>%
  mutate(main_type = str_sub(type, 1, 4)) %>%
  group_by(version, main_type, area_class, sac) %>%
  summarise(area_ha_sac = sum(area_ha),
            n_watersurfaces_sac = sum(n_watersurfaces)) %>%
  ungroup() %>%
  group_by(version, main_type, area_class) %>%
  mutate(n_watersurfaces = sum(n_watersurfaces_sac),
         area_ha = sum(area_ha_sac)) %>%
  ungroup() %>%
  mutate(weight1 = log(area_ha + 1),
          weight2 = area_ha ^ (1/2)) %>%
  group_by(version, main_type, sac) %>%
  mutate(total_area_ha = sum(area_ha * (area_class != "area >= 50 ha")),
         total_watersurfaces = sum(n_watersurfaces * (area_class != "area >= 50 ha")),
         total_weight1 = sum(weight1 * (area_class != "area >= 50 ha")),
         total_weight2 = sum(weight2 * (area_class != "area >= 50 ha"))) %>%
  ungroup() %>%
  mutate(n_finite_fland = total_watersurfaces * n_fl / (n_fl + (total_watersurfaces - 1))) %>%
  mutate(samplesize_fl_alt1 = ifelse(area_class != "area >= 50 ha", n_finite_fland * weight1 / total_weight1, n_watersurfaces),
         samplesize_fl_alt2 = ifelse(area_class != "area >= 50 ha", n_finite_fland * weight2 / total_weight2, n_watersurfaces),
         samplesize_fl_alt3 = ifelse(area_class != "area >= 50 ha", n_finite_fland /3, n_watersurfaces)) %>%
  mutate(samplesize_fl_alt1 = pmin(samplesize_fl_alt1 * n_watersurfaces_sac/n_watersurfaces, n_watersurfaces),
         samplesize_fl_alt2 = pmin(samplesize_fl_alt2 * n_watersurfaces_sac/n_watersurfaces, n_watersurfaces),
         samplesize_fl_alt3 = pmin(samplesize_fl_alt3 * n_watersurfaces_sac/n_watersurfaces, n_watersurfaces)) 


samplesize_sac <- watersurfaces_hab_distribution %>%
  filter(sac) %>%
  mutate(main_type = str_sub(type, 1, 4)) %>%
  group_by(version, sac, main_type, area_class) %>%
  summarise(area_ha = sum(area_ha),
            n_watersurfaces = sum(n_watersurfaces)) %>%
  ungroup() %>%
  mutate(weight1 = log(area_ha + 1),
          weight2 = area_ha ^ (1/2)) %>%
  group_by(version, main_type) %>%
  mutate(total_area_ha = sum(area_ha * (area_class != "area >= 50 ha")),
         total_watersurfaces = sum(n_watersurfaces),
         total_weight1 = sum(weight1 * (area_class != "area >= 50 ha")),
         total_weight2 = sum(weight2 * (area_class != "area >= 50 ha"))) %>%
  ungroup() %>%
  mutate(n_finite_sac = total_watersurfaces * n_sac / (n_sac + (total_watersurfaces - 1))) %>%
  mutate(samplesize_sac_alt1 = ifelse(area_class != "area >= 50 ha", n_finite_sac * weight1 / total_weight1, n_watersurfaces),
         samplesize_sac_alt2 = ifelse(area_class != "area >= 50 ha", n_finite_sac * weight2 / total_weight2, n_watersurfaces),
         samplesize_sac_alt3 = ifelse(area_class != "area >= 50 ha", n_finite_sac /3, n_watersurfaces)) %>%
  mutate(samplesize_sac_alt1 = pmin(samplesize_sac_alt1, n_watersurfaces),
         samplesize_sac_alt2 = pmin(samplesize_sac_alt2, n_watersurfaces),
         samplesize_sac_alt3 = pmin(samplesize_sac_alt3, n_watersurfaces))

samplesize_subtype <- watersurfaces_hab_distribution %>%
  mutate(main_type = str_sub(type, 1, 4)) %>%
  filter(main_type == "3130") %>%
  group_by(version, type, area_class, sac) %>%
  summarise(n_watersurfaces_sac = sum(n_watersurfaces),
            area_ha_sac = sum(area_ha)) %>%
  ungroup() %>%
  group_by(version, type, area_class) %>%
  mutate(n_watersurfaces = sum(n_watersurfaces_sac),
         area_ha = sum(area_ha_sac)) %>%
  ungroup() %>%
  mutate(weight1 = log(area_ha + 1),
          weight2 = area_ha ^ (1/2)) %>%
  group_by(version, type, sac) %>%
  mutate(total_area_ha = sum(area_ha * (area_class != "area >= 50 ha")),
         total_watersurfaces = sum(n_watersurfaces),
         total_weight1 = sum(weight1 * (area_class != "area >= 50 ha")),
         total_weight2 = sum(weight2 * (area_class != "area >= 50 ha"))) %>%
  ungroup() %>%
  mutate(n_finite_subtype = total_watersurfaces * n_subtype / (n_subtype + (total_watersurfaces - 1))) %>%
  mutate(samplesize_subtype_alt1 = ifelse(area_class != "area >= 50 ha", n_finite_subtype * weight1 / total_weight1, n_watersurfaces),
         samplesize_subtype_alt2 = ifelse(area_class != "area >= 50 ha", n_finite_subtype * weight2 / total_weight2, n_watersurfaces),
         samplesize_subtype_alt3 = ifelse(area_class != "area >= 50 ha", n_finite_subtype /3, n_watersurfaces)) %>%
  mutate(samplesize_subtype_alt1 = pmin(samplesize_subtype_alt1 * n_watersurfaces_sac/n_watersurfaces, n_watersurfaces_sac),
         samplesize_subtype_alt2 = pmin(samplesize_subtype_alt2 * n_watersurfaces_sac/n_watersurfaces, n_watersurfaces_sac),
         samplesize_subtype_alt3 = pmin(samplesize_subtype_alt3 * n_watersurfaces_sac/n_watersurfaces, n_watersurfaces_sac))

samplesize_alt <- watersurfaces_hab_distribution %>%
  mutate(main_type = str_sub(type, 1, 4),
         n_fl = n_fl,
         n_sac = n_sac,
         n_subtype = n_subtype) %>%
  select(version, n_fl, n_sac, n_subtype, main_type, type, sac, area_class, area_ha, n_watersurfaces) %>%
  group_by(version, main_type) %>%
  mutate(n_watersurfaces_sample = sum(n_watersurfaces)) %>%
  ungroup() %>%
  left_join(select(samplesize_fl, sac, version, main_type, area_class, samplesize_fl_alt1, samplesize_fl_alt2, samplesize_fl_alt3), by = c("version", "main_type", "area_class", "sac")) %>%
  left_join(select(samplesize_sac, sac, version, main_type, area_class, samplesize_sac_alt1, samplesize_sac_alt2, samplesize_sac_alt3), by = c("version", "main_type", "area_class", "sac")) %>%
  group_by(version, main_type, sac, area_class) %>%
  mutate(n_watersurfaces_main_type = sum(n_watersurfaces)) %>%
  ungroup() %>%
  mutate(samplesize_main_type_alt1 = ifelse(area_class == "area >= 50 ha" | n_watersurfaces_sample < 30, n_watersurfaces,
                                            ifelse(sac, pmax(samplesize_sac_alt1, samplesize_fl_alt1), samplesize_fl_alt1)),
         samplesize_main_type_alt2 = ifelse(area_class == "area >= 50 ha" | n_watersurfaces_sample < 30, n_watersurfaces,
                                            ifelse(sac, pmax(samplesize_sac_alt2, samplesize_fl_alt2), samplesize_fl_alt2)),
         samplesize_main_type_alt3 = ifelse(area_class == "area >= 50 ha" | n_watersurfaces_sample < 30, n_watersurfaces,
                                            ifelse(sac, pmax(samplesize_sac_alt3, samplesize_fl_alt3), samplesize_fl_alt3)),
         samplesize_type_alt1 = samplesize_main_type_alt1 * n_watersurfaces / n_watersurfaces_main_type,
         samplesize_type_alt2 = samplesize_main_type_alt2 * n_watersurfaces / n_watersurfaces_main_type,
         samplesize_type_alt3 = samplesize_main_type_alt3 * n_watersurfaces / n_watersurfaces_main_type,
         ) %>%
  left_join(select(samplesize_subtype, sac, version, type, area_class, samplesize_subtype_alt1, samplesize_subtype_alt2, samplesize_subtype_alt3), by = c("version", "type", "area_class", "sac")) %>%
  mutate(samplesize_total_alt1 = round(pmax(samplesize_type_alt1, samplesize_subtype_alt1, na.rm = TRUE)),
         samplesize_total_alt2 = round(pmax(samplesize_type_alt2, samplesize_subtype_alt2, na.rm = TRUE)),
         samplesize_total_alt3 = round(pmax(samplesize_type_alt3, samplesize_subtype_alt3, na.rm = TRUE)))

  return(samplesize_alt)

}
  

samplesize_alt1 <- samplesize_watersurfaces(watersurfaces_hab_distribution = watersurfaces_hab_distribution,
                                            n_fl = 80, n_sac = 80, n_subtype = 40) %>%
  mutate(scenario = "alternative 1")

samplesize_alt2 <- samplesize_watersurfaces(watersurfaces_hab_distribution = watersurfaces_hab_distribution,
                                            n_fl = 80, n_sac = 0, n_subtype = 40) %>%
  mutate(scenario = "alternative 2")


```

```{r}
samplesize_alt_total <- samplesize_alt1 %>%
  bind_rows(samplesize_alt2) %>%
  select(version, scenario, n_fl, n_sac, n_subtype, type, area_class, sac, n_watersurfaces, samplesize = samplesize_total_alt3) %>% 
  group_by(scenario, n_fl, n_sac, n_subtype, version, area_class) %>%
  summarise(samplesize = sum(samplesize),
            n_watersurfaces = sum(n_watersurfaces)) %>%
  ungroup() %>%
  mutate(samplesize_show = str_c(round(samplesize), " of ", n_watersurfaces),
         .keep = "unused") %>%
  spread(key = area_class, value = samplesize_show, fill = 0) %>%
  bind_rows(samplesize_total)
```

Table \@ref(tab:alttotalws) shows the total sample sizes per area class for the different alternative scenarios. Table \@ref(tab:alttypews) shows the sample sizes per type, per area class and inside/outside SAC for the different alternative scenarios.

```{r alttotalws, eval = TRUE}
samplesize_alt_total %>%
  filter(version == "watersurfaces_hab_2021" | (version == "habitatmap_2014" & scenario == "original")) %>%
  filter(!(version == "watersurfaces_hab_2021" & scenario == "original")) %>%
  select(-version) %>%
  rename("$n_{fl}$" = n_fl, "$n_{sac}$" = "n_sac", "$n_{subtype}$" = "n_subtype" ) %>%
  kable(caption = "Alternative sample size scenarios compared the to original design: total number of sampling units and total number of water surfaces",
        booktabs = TRUE,
        escape = FALSE, 
        format = format_table) %>%
  kable_styling() %>%
  column_spec(c(4,5,6,7), width = "1.8cm")
```

```{r alttypews}
samplesize_alt_overview <- samplesize_alt1 %>%
  bind_rows(samplesize_alt2) %>%
  select(version, scenario, n_fl, n_sac, n_subtype, type, area_class, sac, n_watersurfaces, samplesize = samplesize_total_alt3) %>%
  filter(version %in% c("watersurfaces_hab_2021")) %>%
  # filter(type != "2190_a") %>%
  mutate(samplesize_show = str_c(round(samplesize), " of ", n_watersurfaces),
         .keep = "unused") %>%
  spread(key = area_class, value = samplesize_show, fill = 0) %>%
  select(type, version, scenario, everything()) 

samplesize_alt_overview %>%
  arrange(type, scenario) %>%
  select(-version, -n_fl, -n_sac, -n_subtype) %>%
  mutate(sac = ifelse(sac, "inside", "outside")) %>%
  kable(caption = "Alternative sample size scenarios compared the to original design: number of sampling units per type and total number of water surfaces per type",
        booktabs = TRUE, 
        format = format_table) %>%
  kable_styling() %>%
  column_spec(c(4,5,6,7), width = "1.8cm") %>%
  collapse_rows(c(1,2), target = 1)
  
```

Sample sizes of the alternative scenarios are slightly higher compared to the original sample size (except for 3130_aom). However, this increase is smaller compared to the case where the original sample size calculation is applied to watersurfaces_hab.

We choose alternative 2 (no extra sampling units within sac) as most of the water bodies with habitat are located within SAC, and therefore a sample on the scale of Flanders already contains sufficient sampling units within SAC.

In Table \@ref(tab:tab-wsyear) we compare the number of water bodies that have to be measured per year in the original design and the updated design. Since measuring large water bodies requires more effort, we show the numbers per area class. We see that in the updated design a slightly higher number of large water bodies have to be measured.

```{r tab-wsyear, eval =TRUE}

total <- samplesize %>%
  filter(version == "habitatmap_2014") %>%
  mutate(scenario = "original design") %>%
  group_by(scenario, area_class) %>%
  summarise(n = ceiling(sum(samplesize))) %>%
  ungroup()

total_alt <- samplesize_alt2 %>%
  select(version, scenario, n_fl, n_sac, n_subtype, type, area_class, sac, n_watersurfaces, samplesize = samplesize_total_alt3) %>%
  filter(version %in% c("watersurfaces_hab_2021")) %>%
  mutate(scenario = "updated design") %>%
  group_by(scenario, area_class) %>%
  summarise(n = ceiling(sum(samplesize))) %>%
  ungroup()

total %>%
  bind_rows(total_alt) %>%
  mutate(n_year = round(n/6, 1)) %>%
  select(-n) %>%
  spread(key = area_class, value = n_year) %>%
  kable(booktabs = TRUE,
        caption = "Number of sampling unit per year for the different area classes in both the original design and the updated design", 
        format = format_table) %>%
  kable_styling()
           
  
```
The first monitoring cycle will be finished in 2022 and the new monitoring design will be implemented in the second monitoring cycle starting from 2023. 

## Sampling process

As for the terrestrial habitat types, the sampling process relies on the grts-master sample for Flanders [@onkelinx2019]. To assign a grts-ranking to a water body (the population unit) we select a reference point in the water body using the function `st_point_on_surface` of the sf package [@Pebesma2018]. Next, the water body gets the grts-ranking of the grts-master grid cell in which the reference point is located. Once a grts-ranking is assigned to all water bodies, a sample can be drawn in a similar way as the terrestrial habitat types. 










